---
title: "JASA_simulation 1"
author: "Ziyi Wang"
date: "2026-02-07"
output: html_document
---
## Simulation 1 - One dimensional x 
sim a


### Load Packages/Functions

```{r setup, include=FALSE}
library(knitr)
library(locfdr)
library(CATMicrobiome)
library(admix)
library(mvtnorm)
library(mixtools)
library(devtools)
library(KScorrect)
library(mclust)
library(adaptMT)
library(AdaPTGMM)
library(splines2)
library(splines)
library(dplyr)
library(ggplot2)
library(tidyr)
library(coin)
library(ggrepel)
library(viridisLite)
library(IHW)
library(FDRreg)
library(OrderShapeEM)
library(chai)

# setwd("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian")

# ggsave("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/plots/Simulations/SimA_fixa_1_2.png", plot = a_1_2, width = 14, height = 9, units = "in", dpi = 300)

# devtools::load_all("C:/Ziyi Wang/chai_package/chai")
```

### Simulation generating and fit the models

Note: this chunk is very time-consuming. Specifically, for each seed (100 randomly generated seeds in total) and each value of $a$ ($a \in {0, 0.5, 1, 1.5, 2}$), one simulation dataset was generated and then analyzed using \code{chai} and each benchmark method. For every method, we recorded the number of selections, FDP, and power at all $q$-level from 0.01 to 0.10, together with the corresponding seed and $a$ setting. 

You can directly run next chuck to load all the results.

```{r, eval=FALSE}
run_all_methods_sim <- function(seed, a_val, q_grid, 
                                n0 = 950,  n1 = 25,   n2 = 25,
                                m0 = 0,    m1 = 2,    m2 = -4,
                                sd0 = 1,   sd1 = 1,   sd2 = 1){
  set.seed(seed)
  
  # Simulation generation - z
  n <- n0 + n1 + n2
  z0 <- rnorm(n0, mean = 0, sd = 1)
  z1 <- rnorm(n1, mean = 2, sd = 0.5)
  z2 <- rnorm(n2, mean = -2, sd = 0.5)
  z  <- c(z0, z1, z2)
  # gt <- seq((n0 + 1), n)
  p <- 2 * (1 - pnorm(abs(z)))
  
  # Simulation generation - X
  m0a <- m0
  m1a <- m0 + a_val * (m1 - m0)
  m2a <- m0 + a_val * (m2 - m0)
  
  x0 <- rnorm(n0, mean = m0a, sd = sd0)
  x1 <- rnorm(n1, mean = m1a, sd = sd1)
  x2 <- rnorm(n2, mean = m2a, sd = sd2)
  X  <- c(x0, x1, x2)
  x_df <- data.frame(X = X)
  formula <- paste0("ns(X, df = ", 2:8, ")")
  
  # Initialize a list
  run_results <- list()
  
  ## 1) chai 
  res_chai <- chai(z, X, B = 100)
  
  for (qq in seq_along(q_grid)) {
    select <- clfdrselect(res_chai, q = q_grid[qq])
    p1  <- performance(gt, select, z = res_chai$z,
                       bh_level = q_grid[qq], one.sided = FALSE)
    # chai
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "chai", 
                  Selected = p1[[4]], FDP = p1[[1]], Power = p1[[2]])
    # BH
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "BH", 
                  Selected = p1[[13]], FDP = p1[[8]], Power = p1[[9]])
  }

  ## 2) adapt_gmm (z)
  res_gmm_z <- adapt_gmm(x_df, z = z, alphas = q_grid,
                         beta_formulas = formula, testing = "two_sided")
  
  for (qq in seq_along(q_grid)) {
    p2  <- performance(gt, res_gmm_z$rejs[[qq]], z = z,
                       bh_level = q_grid[qq], one.sided = FALSE)
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "adapt_gmm_z", 
                  Selected = p2[[4]], FDP = p2[[1]], Power = p2[[2]])
  }

  ## 3) adapt_gmm (p)
  res_gmm_p <- adapt_gmm(x_df, pvals = p, alphas = q_grid,
                         beta_formulas = formula, testing = "two_sided")
  for (qq in seq_along(q_grid)) {
    # idx <- match(qq, q_grid)
    p3  <- performance(gt, res_gmm_p$rejs[[qq]], z = z,
                       bh_level = q_grid[qq], one.sided = FALSE)
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "adapt_gmm_p", 
                  Selected = p3[[4]], FDP = p3[[1]], Power = p3[[2]])
  }

  ## 4) adapt_glm
  res_glm <- adapt_glm(x_df, pvals = p, alphas = q_grid,
                       pi_formulas = formula, mu_formulas = formula)
  empty <- is.null(res_glm) || (!is.list(res_glm$rejs))
  for (qq in seq_along(q_grid)) {
    if (!empty && lengths(res_glm$rejs)[qq] > 0L){
     p4 <- performance(gt, res_glm$rejs[[qq]], z = z,
                        bh_level = q_grid[qq], one.sided = FALSE)
     run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "adapt_glm", 
                  Selected = p4[[4]], FDP = p4[[1]], Power = p4[[2]])
    } else {
      run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "adapt_glm", 
                  Selected = 0, FDP = NA, Power = NA)
      }
  }

  ## 5) OrderShapeEM
  tryCatch({
    res_osem <- OrderShapeEM(pvals = p, order.var = X,
                             OrderShapeEM.control(trace = FALSE))
    for (qq in seq_along(q_grid)) {
      p5 <- performance(gt, which(res_osem$fdr <= q_grid[qq]), z = z,
                        bh_level = q_grid[qq], one.sided = FALSE)
      run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "OrderShapeEM", 
                  Selected = p5[[4]], FDP = p5[[1]], Power = p5[[2]])
    }
  }, error = function(e) {
    message(paste("OrderShapeEM failed at seed", seeds[s], "a =", a, ":", e$message))
    for (qq in seq_along(q_grid)) {
      run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "OrderShapeEM", 
                  Selected = NA, FDP = NA, Power = NA)
    }
  })

  ## 6) FDRreg（theoretical null）
  res_fdrreg <- FDRreg(z, as.matrix(X), nulltype = "theoretical")
  for (qq in seq_along(q_grid)) {
    p6 <- performance(gt, which(res_fdrreg$FDR <= q_grid[qq]), z = z,
                      bh_level = q_grid[qq], one.sided = FALSE)
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "FDRreg", 
                  Selected = p6[[4]], FDP = p6[[1]], Power = p6[[2]])
  }
    
    
  ## 7) IHW
  for (qq in seq_along(q_grid)) {
    res_ihw <- IHW::ihw(pvalues = p, covariates = x_df[,1], alpha = q_grid[qq], nbins = 5)
    p7 <- performance(gt, which(IHW::adj_pvalues(res_ihw) <= q_grid[qq]), z = z, bh_level = qq, one.sided = FALSE)
    
    run_results[[length(run_results)+1]] <- 
       data.frame(Seed = seed, a = a_val, 
                  q = q_grid[qq], Method = "IHW", 
                  Selected = p7[[4]], FDP = p7[[1]], Power = p7[[2]])
  }
  
  # Clean up memory after this specific a_val/seed is done
  gc()
  
  return(do.call(rbind, run_results))
}

set.seed(1)
seeds <- c(sample(1:10000, 99), 123)  # Add seed 123 for checking purpose

n  <- 1000
n0 <- 950
n1 <- 25
m0 <- 0; m1 <- 2; m2 <- -4
sd0 <- 1;   sd1 <- 1; sd2 <- 1

# ground truth
gt <- seq((n0 + 1), n)
  
# informativeness grid：from 'no info' to 'higher info'
a_grid <- c(0, 0.5, 1, 1.5, 2)
q_grid <- seq(0.01, 0.1, by = 0.01)

cache_file <- "all_methods_simulation_1_results.rds"

if (file.exists(cache_file)){
  # if already did, load it
  results_all <- readRDS(cache_file)
} else {
  # if not, run the simulation
  param_grid <- expand.grid(seed = seeds, a = a_grid)
  
  library(purrr)
  results_list <- pmap(list(param_grid$seed, param_grid$a),
                       ~run_all_methods_sim(seed = ..1, a_val = ..2, q_grid = q_grid))
  
  results_all <- bind_rows(results_list) %>% mutate(across(c(Selected, FDP, Power), as.numeric))
  
  # save to dist
  saveRDS(results_all, cache_file)
}
```

### Load simulation 1's results 
Load simulation 1 results and calculate the average across 100 repeats.

```{r}
# Load results
results_1 <- readRDS("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/chai_env/simA_all_results_with_zero.rds")

# # Load summary results
# summary_1 <- readRDS("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/chai_env/simA_summary_with_zero.rds")

# Calculate the average of selections, FDP and power.
summary_1 <- results_1 %>%
  group_by(Method, q, a) %>%
  summarise(
    mean_Selected = mean(Selected),
    mean_FDP      = mean(FDP),
    mean_Power    = mean(Power),
    .groups = "drop"
  )

# Check the summary results at a = 1 and q = 0.05
summary_1 %>% filter(a == 1, q == 0.05)

simA_long <- summary_1 %>%
  mutate(
    mean_FDP   = ifelse(is.nan(mean_FDP), NA, mean_FDP),
    mean_Power = ifelse(is.nan(mean_Power), NA, mean_Power)
  ) %>%
  pivot_longer(cols = c(mean_FDP, mean_Power),
               names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = recode(Metric,
                         mean_FDP = "FDP",
                         mean_Power = "Power"))
```

### Define the plot functions to ensure consistent color assignments across different methods.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# ---------- color helpers ----------
blend_color <- function(col, target = "white", amount = 0.4) {
  a <- grDevices::col2rgb(col) / 255
  b <- grDevices::col2rgb(target) / 255
  out <- (1 - amount) * a + amount * b
  grDevices::rgb(out[1,], out[2,], out[3,])
}

variant_shades <- function(base_col, K = 12) {
  grDevices::colorRampPalette(c(
    blend_color(base_col, "white", 0.55),
    base_col,
    blend_color(base_col, "black", 0.25)
  ))(K)
}

make_method_palette <- function(methods, K = 12) {
  fixed <- c(
    chai        = "red",
    adapt_glm   = "#FF7F0E",
    adapt_gmm_z = "#2CA02C",
    adapt_gmm_p = "#9467BD",
    FDRreg      = "#8C564B",
    BH          = "black"
  )

  family_base <- c(
    OrderShapeEM = "#1F77B4",  # blue family
    IHW          = "#17BECF"   # teal family
  )

  ord_shades <- variant_shades(family_base["OrderShapeEM"], K = K)
  ihw_shades <- variant_shades(family_base["IHW"],          K = K)

  methods <- unique(as.character(methods))
  pal <- setNames(rep("#7F7F7F", length(methods)), methods)  # fallback grey

  for (m in methods) {
    if (m %in% names(fixed)) {
      pal[m] <- fixed[m]
      next
    }

    base <- sub("_x\\d+$", "", m)
    idx  <- suppressWarnings(as.integer(sub(".*_x", "", m)))

    if (base %in% names(family_base)) {
      if (is.na(idx)) {
        pal[m] <- family_base[base]  # e.g. "OrderShapeEM" with no suffix
      } else {
        idx <- max(1L, min(idx, K))
        pal[m] <- if (base == "OrderShapeEM") ord_shades[idx] else ihw_shades[idx]
      }
    }
  }

  pal
}

method_order <- function(methods) {
  methods <- unique(as.character(methods))
  core <- c("chai","adapt_glm","adapt_gmm_z","adapt_gmm_p","FDRreg")

  sort_family <- function(base) {
    fam <- methods[grepl(paste0("^", base, "(_x\\d+)?$"), methods)]
    base_present <- base %in% fam
    vars <- fam[grepl("_x\\d+$", fam)]
    if (length(vars)) {
      idx <- suppressWarnings(as.integer(sub(".*_x", "", vars)))
      vars <- vars[order(idx)]
    }
    c(if (base_present) base else character(0), vars)
  }

  out <- c(core, sort_family("OrderShapeEM"), sort_family("IHW"), "BH")
  out <- unique(out)
  out[out %in% methods]
}

# ---------- Updated plotting function (fix a or q) ----------
library(ggrepel)
library(stringr)

plot_metric_aq <- function(long_table, metric_name, x_var = c("q","a"),
                           ylab, title,
                           K_variants = 12,
                           fdp_y = 0.2,
                           x_breaks = NULL,
                           q_level = 0.05,
                           method_font = 5.5) {

  x_var <- match.arg(x_var)

  dat <- long_table %>%
    dplyr::filter(Metric == metric_name) %>%
    mutate(Method = as.character(Method))

  ord <- method_order(dat$Method)
  dat <- dat %>% mutate(Method = factor(Method, levels = ord))
  pal <- make_method_palette(levels(dat$Method), K = K_variants)

  # endpoints for labels = last x (either max q or max a)
  endpoints <- dat %>%
    group_by(Method) %>%
    filter(.data[[x_var]] == max(.data[[x_var]], na.rm = TRUE)) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    mutate(label = as.character(Method))
  
  y_max <- if (metric_name == "FDP") fdp_y else 1
  
  endpoints <- endpoints %>%
  mutate(
    Value_label = pmin(Value, y_max)   # clamp label anchor to the axis top
  )

  x_rng   <- range(dat[[x_var]], na.rm = TRUE)
  x_nudge <- diff(x_rng) * 0.03

  p <- ggplot(dat, aes(x = .data[[x_var]], y = Value, group = Method, color = Method)) +
    geom_line(linewidth = 0.9, na.rm = TRUE) +
    geom_point(size = 1.6, na.rm = TRUE) +
    ggrepel::geom_text_repel(
      data = endpoints,
      aes(x = .data[[x_var]], y = Value_label, label = label),
      size = method_font,
      nudge_x = x_nudge,
      hjust = 0,
      direction = "y",
      box.padding = 0.25,
      point.padding = 0.15,
      segment.size = 0.25,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 1,
      show.legend = FALSE
    ) +
    scale_color_manual(values = pal, breaks = ord) +
    coord_cartesian(ylim = c(0, 1), clip = "off") +
    labs(x = x_var, y = ylab, title = title) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.margin = margin(5.5, 60, 5.5, 5.5),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      axis.text.x  = element_text(size = 12),
      axis.text.y  = element_text(size = 12)
    )

  if (!is.null(x_breaks)) {
    p <- p + scale_x_continuous(breaks = x_breaks,
                                expand = expansion(mult = c(0.02, 0.18)))
  } else {
    p <- p + scale_x_continuous(expand = expansion(mult = c(0.02, 0.18)))
  }

  if (metric_name == "FDP" && x_var == "q") {
    p <- p +
      annotate(
        "segment",
        x = 0, y = 0, xend = 0.1, yend = 0.1,
        linetype = "dashed",
        linewidth = 1.2,
        color = "black"
      ) +
      coord_cartesian(ylim = c(0, fdp_y))

    
      # geom_segment(
      #   aes(x = 0, y = 0, xend = 0.1, yend = 0.1),
      #   inherit.aes = FALSE,
      #   linetype = "dashed",
      #   linewidth = 1.2,
      #   color = "black"
      # ) +
  }
  
  if (metric_name == "FDP" && x_var == "a") {
    p <- p +
      geom_hline(
        yintercept = q_level,
        linetype = "dashed",
        linewidth = 1.2,
        color = "black"
      ) +
      coord_cartesian(ylim = c(0, fdp_y))
  }

  p
}

```

### Visualization - Fix a

```{r}
library(patchwork)

a0_dat <- dplyr::filter(simA_long, abs(a - 0) < 1e-12)
a0_fdp <- plot_metric_aq(a0_dat, "FDP", x_var = "q",
                         ylab = "Mean FDP",
                         title = sprintf("FDP vs q (a = 0)"),
                         fdp_y = 0.12,
                         x_breaks = sort(unique(a0_dat$q))) 
a0_pow <- plot_metric_aq(a0_dat, "Power", x_var = "q",
                            ylab = "Mean Power",
                            title = sprintf("Power vs q (a = 0)"),
                            x_breaks = sort(unique(a0_dat$q)))

a0.5_dat <- dplyr::filter(simA_long, abs(a - 0.5) < 1e-12)
a0.5_fdp <- plot_metric_aq(a0.5_dat, "FDP", x_var = "q",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs q (a = 0.5)"),
                            fdp_y = 0.12,
                            x_breaks = sort(unique(a0.5_dat$q)))
a0.5_pow <- plot_metric_aq(a0.5_dat, "Power", x_var = "q",
                            ylab = "Mean Power",
                            title = sprintf("Power vs q (a = 0.5)"),
                            x_breaks = sort(unique(a0.5_dat$q)))

a1_dat <- dplyr::filter(simA_long, abs(a - 1) < 1e-12)
a1_fdp <- plot_metric_aq(a1_dat, "FDP", x_var = "q",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs q (a = 1)"),
                            fdp_y = 0.12,
                            x_breaks = sort(unique(a1_dat$q)))
a1_pow <- plot_metric_aq(a1_dat, "Power", x_var = "q",
                            ylab = "Mean Power",
                            title = sprintf("Power vs q (a = 1)"),
                            x_breaks = sort(unique(a1_dat$q)))


a1.5_dat <- dplyr::filter(simA_long, abs(a - 1.5) < 1e-12)
a1.5_fdp <- plot_metric_aq(a1.5_dat, "FDP", x_var = "q",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs q (a = 1.5)"),
                            fdp_y = 0.12,
                            x_breaks = sort(unique(a1.5_dat$q)))
a1.5_pow <- plot_metric_aq(a1.5_dat, "Power", x_var = "q",
                            ylab = "Mean Power",
                            title = sprintf("Power vs q (a = 1.5)"),
                            x_breaks = sort(unique(a1.5_dat$q)))

a2_dat <- dplyr::filter(simA_long, abs(a - 2) < 1e-12)
a2_fdp <- plot_metric_aq(a2_dat, "FDP", x_var = "q",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs q (a = 2)"),
                            fdp_y = 0.12,
                            x_breaks = sort(unique(a2_dat$q)))
a2_pow <- plot_metric_aq(a2_dat, "Power", x_var = "q",
                            ylab = "Mean Power",
                            title = sprintf("Power vs q (a = 2)"),
                            x_breaks = sort(unique(a2_dat$q)))


fdps  <- list(a0_fdp, a0.5_fdp, a1_fdp, a1.5_fdp, a2_fdp)
pows  <- list(a0_pow, a0.5_pow, a1_pow, a1.5_pow, a2_pow)

combo <- wrap_plots(
  c(fdps, pows),
  nrow = 5, ncol = 2,
  byrow = FALSE
) 
  # + plot_annotation(title = "Method Performance vs q")

# combo

a_1_2 <- wrap_plots(
  c(list(a1_fdp, a2_fdp), list(a1_pow, a2_pow)),
  nrow = 2, ncol = 2,
  byrow = FALSE
)
  # + plot_annotation(title = "Method Performance vs q")
# a_1_2

# ggsave("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/plots/Simulations/SimA_fixa_fdr_power_updated_zero.png", plot = combo, width = 14, height = 18, units = "in", dpi = 300)
# 
# ggsave("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/plots/Simulations/SimA_fixa_1_2_zero.png", plot = a_1_2, width = 14, height = 9, units = "in", dpi = 300)

```

```{r, fig.width=14, fig.height=18, out.width='100%', warning=FALSE}
combo
```


```{r, fig.width=14, fig.height=9, out.width='100%', warning=FALSE}
a_1_2
```


### Visualization - Fix q

```{r}
library(patchwork)

q0.01_dat <- dplyr::filter(simA_long, abs(q - 0.01) < 1e-12)
q0.01_fdp <- plot_metric_aq(q0.01_dat, "FDP", x_var = "a",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs a (q = 0.01)"),
                            fdp_y = 0.05,
                            x_breaks = sort(unique(q0.01_dat$a)),
                            q_level = 0.01)

q0.01_pow <- plot_metric_aq(q0.01_dat, "Power", x_var = "a",
                            ylab = "Mean Power",
                            title = sprintf("Power vs a (q = 0.01)"),
                            x_breaks = sort(unique(q0.01_dat$a)),
                            q_level = 0.01)


q0.05_dat <- dplyr::filter(simA_long, abs(q - 0.05) < 1e-12)
q0.05_fdp <- plot_metric_aq(q0.05_dat, "FDP", x_var = "a",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs a (q = 0.05)"),
                            fdp_y = 0.1,
                            x_breaks = sort(unique(q0.05_dat$a)),
                            q_level = 0.05)

q0.05_pow <- plot_metric_aq(q0.05_dat, "Power", x_var = "a",
                            ylab = "Mean Power",
                            title = sprintf("Power vs a (q = 0.05)"),
                            x_breaks = sort(unique(q0.05_dat$a)),
                            q_level = 0.05)



q0.1_dat <- dplyr::filter(simA_long, abs(q - 0.1) < 1e-12)
q0.1_fdp <- plot_metric_aq(q0.1_dat, "FDP", x_var = "a",
                            ylab = "Mean FDP",
                            title = sprintf("FDP vs a (q = 0.1)"),
                            fdp_y = 0.2,
                            x_breaks = sort(unique(q0.1_dat$a)),
                            q_level = 0.1)

q0.1_pow <- plot_metric_aq(q0.1_dat, "Power", x_var = "a",
                            ylab = "Mean Power",
                            title = sprintf("Power vs a (q = 0.1)"),
                            x_breaks = sort(unique(q0.1_dat$a)),
                            q_level = 0.1)



fdps_pow  <- list(q0.01_fdp, q0.05_fdp, q0.1_fdp)
pows_pow  <- list(q0.01_pow, q0.05_pow, q0.1_pow)

combo_q <- wrap_plots(
  c(fdps_pow, pows_pow),
  nrow = 3, ncol = 2,
  byrow = FALSE
) 
  # + plot_annotation(title = "Method Performance vs a")

# ggsave("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian/plots/Simulations/SimA_fixq_fdr_power_updated_zero.png", plot = combo_q, width = 14, height = 12, units = "in", dpi = 300)

```

```{r, fig.width=14, fig.height=12, out.width='100%', warning=FALSE}
combo_q
```