---
title: "Realdata_Schizophrenia"
author: "Ziyi Wang"
date: "2026-03-16"
output: html_document
---
# Real data
## Schizophrenia - The third real data

### Load Packages/Functions
```{r setup, include=TRUE, warning=FALSE}
suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(SummarizedExperiment)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(GGally)
  library(knitr)
  library(mvtnorm)
  library(mixtools)
  library(devtools)
  library(KScorrect)
  library(adaptMT)
  library(AdaPTGMM)
  library(splines2)
  library(ggplot2)
  library(coin)
  library(ggrepel)
  library(IHW)
  library(FDRreg)
  library(OrderShapeEM)
  library(chai)
})
```

```{r, include=FALSE}
# Set the working direction
setwd("C:/Users/zwang26/OneDrive - UTHealth Houston/conditionalGaussian")
```

### Load the data and pre-processing the data

**Note**: The pro-processing part and model fitting chunk could be time-consuming. We saved the processed data together with the organized results after fitting the models. When knitting this HTML file, we loaded the **Schizophrenia_env.RData** file directly and generated the plots from it. 

However, for verification or reproducibility, you can also run each pre-processing chunk and/or model-fitting chunk one by one, while skiping the chunk that loads the .RData file. This should produce the same results.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
ae5  <- read.csv("./autoencoder_project/outputs/ZhuF_2020_species_AE5.csv", row.names = 1)
pca5 <- read.csv("./autoencoder_project/outputs/ZhuF_2020_species_PCA5.csv", row.names = 1) 

# Helper: extract species label
.extract_species <- function(taxon) {
  sp <- ifelse(grepl("s__", taxon, fixed = TRUE),
               sub(".*s__([^|;]+).*", "\\1", taxon),
               taxon)
  sp <- gsub("^s__", "", sp)
  sp <- gsub("_", " ", sp)
  trimws(sp)
}

# Extract the relative abundance
abun <- curatedMetagenomicData("ZhuF_2020.relative_abundance", dryrun = FALSE)

# Extract the assay matrix
abun_mat <- assay(abun[[1]], "relative_abundance")
# 480 x 171

r_names <- rownames(abun_mat)
taxa_names <- sub("^.*\\|", "", r_names)
species_clean <- .extract_species(taxa_names)
rownames(abun_mat) <- species_clean
# 480 x 171

length(intersect(rownames(ae5), rownames(abun_mat)))
common_species <- intersect(rownames(ae5), rownames(abun_mat))
# 323 sharing

abun_mat_common <- abun_mat[common_species, , drop = FALSE]

# Column-Normalize (column is sample)
cs <- colSums(abun_mat_common, na.rm = TRUE)
cs[cs == 0] <- 1
abun_mat_norm <- sweep(abun_mat_common, 2, cs, FUN = "/")

t_abun <- t(abun_mat_norm)
# 171 subjects x 323 species

######################### outcome ###############################
outcome <- as.factor(abun$`2021-03-31.ZhuF_2020.relative_abundance`[[5]])
# schizophrenia 90 vs. control 81
```

### Generating the p-value/z-value

```{r, eval=FALSE}
compute_pz_wilcox <- function(t_abun, outcome) {
  stopifnot(nrow(t_abun) == length(outcome))

  p_z_values <- sapply(colnames(t_abun), function(feature) {
    features <- t_abun[, feature, drop = FALSE]
    wilcox <- wilcox_test(features[, 1] ~ outcome)

    c(
      p_value = as.numeric(pvalue(wilcox)),
      z_value = as.numeric(statistic(wilcox))
    )
  }, simplify = "matrix")
  p_z_values <- as.data.frame(t(p_z_values))
  p_z_values$feature <- rownames(p_z_values)
  rownames(p_z_values) <- NULL

  return(p_z_values)
}

p_z_values <- compute_pz_wilcox(t_abun, outcome)

p_value <- p_z_values[,1]
z_value <- p_z_values[,2]
```


### Load the processed data

The processed data included: 
- **z_value**: the z-statistics from wilcoxon text of each genera between gastrectomy vs control
- **p_value**: the p-value calculated from z-statistics with two-sided
- **outcome**: the outcome (gastrectomy or control) of this dataset  
- **t_abun**: the relative abundance table of this schizophrenia dataset 

- **ae5**: the 5-dimensional autoencoder 
- **pca5**: the 5-dimensional PCA
- **noise**: the randomly generated noise vector that following a standard normal distribution $~N(0, 1)$
- **noise2**: the randomly generated noise vector that following a uniform distribution $~U(0, 1)$

- **chai_zhu_ae5**: \code(chai) output with z_value and ae5 as inputs
- **chai_zhu_pca5**: \code(chai) output with z_value and pca5 as inputs
- **chai_zhu_noise**: \code(chai) output with z_value and noise as inputs
- **chai_zhu_noise2**: \code(chai) output with z_value and noise2 as inputs

- **zhu_ae**: the summary table saving the number of selections of every method at different q, with ae5 as side information.
- **zhu_ae_long**: the long table version of zhu_ae for plot purpose.

- **zhu_ae_pca**: the summary table saving the number of selections of every method at different q, with different side information (included ae5, pca5, noise, and noise2).
- **zhu_ae_pca_long**: the long table version of zhu_ae_pca for plot purpose.


```{r}
load("./chai_env/Schizophrenia_env.RData") 
```


### Fit models
```{r, eval=FALSE, message=FALSE, warning=FALSE}
q_levels <- seq(0.01, 0.1, by = 0.01)

################## random X #######################
set.seed(123)
# Standard normal noise: N(0, 1)
noise <- rnorm(length(z_value))

# Uniform noise on U(0, 1)
noise2 <- runif(length(z_value))

################## chai #######################
set.seed(123) 
chai_zhu_ae5 <- chai(z_value, ae5, B = 100)       # autoencoder
chai_rejs_ae5 <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_zhu_ae5$clFDR, q, 1))
})

chai_zhu_pca5 <- chai(z_value, pca5, B = 100)     # pca
chai_rejs_pca5 <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_zhu_pca5$clFDR, q, 1))
})

chai_zhu_noise <- chai(z_value, noise, B = 100)   # noise 1 (~N(0, 1))
chai_rejs_noise <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_zhu_noise$clFDR, q, 1))
})

chai_zhu_noise2 <- chai(z_value, noise2, B = 100) # noise 2 (~U(0, 1))
chai_rejs_noise2 <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_zhu_noise2$clFDR, q, 1))
})

################## Adapt-glm #######################
# autoencoder
x_ae5 <- data.frame(ae5)
make_ns_formulas <- function(dfs) {
  lapply(dfs, function(d)
    as.formula(paste0("~ ns(AE1, df=", d, ") + ns(AE2, df=", d, ") + ns(AE3, df=", d, ") + 
                      ns(AE4, df=", d, ") + ns(AE5, df=", d, ")"))
  )
}
dfs <- c(2, 4, 6, 8)
formula_ns_ae5 <- make_ns_formulas(dfs)

glm_ns_ae5 <- adapt_glm(x_ae5, pvals = p_value, alphas=q_levels, 
                        pi_formulas = formula_ns_ae5, mu_formulas = formula_ns_ae5)
adapt_glm_ns_ae5 <- glm_ns_ae5$nrejs

# pca
x_pca5 <- data.frame(pca5)
make_ns_formulas <- function(dfs) {
  lapply(dfs, function(d)
    as.formula(paste0("~ ns(PC1, df=", d, ") + ns(PC2, df=", d, ") + ns(PC3, df=", d, ") +
                      ns(PC4, df=", d, ") + ns(PC5, df=", d, ")"))
  )
}
dfs <- c(2, 4, 6, 8)
formula_ns_pca5 <- make_ns_formulas(dfs)

glm_ns_pca5 <- adapt_glm(x_pca5, pvals = p_value, alphas=q_levels,
                         pi_formulas = formula_ns_pca5, mu_formulas = formula_ns_pca5)
adapt_glm_ns_pca5 <- glm_ns_pca5$nrejs

# ns - noise
x_noise <- data.frame(noise)
formula_ns_noise <- paste0("ns(noise, df = ", 2:8, ")")

glm_ns_noise <- adapt_glm(x_noise, pvals = p_value, alphas=q_levels,
                          pi_formulas = formula_ns_noise, mu_formulas = formula_ns_noise)
adapt_glm_ns_noise <- glm_ns_noise$nrejs

# ns - noise2
x_noise2 <- data.frame(noise2)
formula_ns_noise2 <- paste0("ns(noise2, df = ", 2:8, ")")

glm_ns_noise2 <- adapt_glm(x_noise2, pvals = p_value, alphas=q_levels,
                           pi_formulas = formula_ns_noise2, mu_formulas = formula_ns_noise2)
adapt_glm_ns_noise2 <- glm_ns_noise2$nrejs

################## Adapt-gmm-p #######################
# autoencoder
gmm_ns_p_ae5 <- adapt_gmm(x_ae5, pvals = p_value,  alphas=q_levels, 
                          beta_formulas = formula_ns_ae5)
adapt_gmm_p_ns_ae5 <- gmm_ns_p_ae5$nrejs

# pca
gmm_ns_p_pca5 <- adapt_gmm(x_pca5, pvals = p_value,  alphas=q_levels,
                           beta_formulas = formula_ns_pca5)
adapt_gmm_p_ns_pca5 <- gmm_ns_p_pca5$nrejs

# ns - noise
gmm_ns_p_noise <- adapt_gmm(x_noise, pvals = p_value,  alphas=q_levels,
                            beta_formulas = formula_ns_noise)
adapt_gmm_p_ns_noise <- gmm_ns_p_noise$nrejs

# ns - noise2
gmm_ns_p_noise2 <- adapt_gmm(x_noise2, pvals = p_value,  alphas=q_levels,
                             beta_formulas = formula_ns_noise2)
adapt_gmm_p_ns_noise2 <- gmm_ns_p_noise2$nrejs


################## Adapt-gmm-z #######################
# autoencoder
gmm_ns_z_ae5 <- adapt_gmm(x_ae5, z = z_value, alphas=q_levels, 
                          beta_formulas = formula_ns_ae5, testing = "two_sided")
adapt_gmm_z_ns_ae5 <- gmm_ns_z_ae5$nrejs

# pca
gmm_ns_z_pca5 <- adapt_gmm(x_pca5, z = z_value, alphas=q_levels,
                           beta_formulas = formula_ns_pca5, testing = "two_sided")
adapt_gmm_z_ns_pca5 <- gmm_ns_z_pca5$nrejs


# ns - noise
gmm_ns_z_noise <- adapt_gmm(x_noise, z = z_value,  alphas=q_levels,
                            beta_formulas = formula_ns_noise, testing = "two_sided")
adapt_gmm_z_ns_noise <- gmm_ns_z_noise$nrejs

# ns - noise2
gmm_ns_z_noise2 <- adapt_gmm(x_noise2, z = z_value,  alphas=q_levels,
                             beta_formulas = formula_ns_noise2, testing = "two_sided")
adapt_gmm_z_ns_noise2 <- gmm_ns_z_noise2$nrejs


################## OrderShapeEM #######################
# Define a function that can fit the easily fit every column of side information into OrderShapeEM
run_ordershape_by_col <- function(mat, dataset = c("ae5", "pca5")) {
  dataset <- match.arg(dataset)
  
  res <- vector("list", ncol(mat))
  names(res) <- paste0("ordershape_x", seq_len(ncol(mat)), "_", dataset)
  
  for (j in seq_len(ncol(mat))) {
    message("Fitting OrderShapeEM for column ", j)
    res[[j]] <- OrderShapeEM(
      pvals     = p_value,
      order.var = mat[, j],
      OrderShapeEM.control(trace = TRUE)
    )
  }
  res
}

# Define a function to count how many selections that OrderShapeEM made at each q level
count_ordershape_selections <- function(os_list, q_levels) {
  # os_list: list like ordershape_ae5 or ordershape_pca5
  # returns a matrix: rows = q_levels, cols = each x1/x2/...
  res <- sapply(os_list, function(os) {
    # for one ordershape object
    sapply(q_levels, function(q) sum(os$fdr <= q))
  })
  # sapply gives q_levels in rows; make that explicit:
  rownames(res) <- paste0("q=", q_levels)
  res
}

# autoencoder
ordershape_ae5 <- run_ordershape_by_col(ae5, dataset = "ae5")
ordershape_ae5  <- count_ordershape_selections(ordershape_ae5,  q_levels)

# pca
ordershape_pca5 <- run_ordershape_by_col(pca5, dataset = "pca5")
ordershape_pca5 <- count_ordershape_selections(ordershape_pca5, q_levels)

# noise
ordershape_noise <- OrderShapeEM(pvals = p_value, order.var = noise,
                                 OrderShapeEM.control(trace = TRUE))
ordershapeem_noise <- sapply(q_levels, function(q) {
  sum(ordershape_noise$fdr <= q)
})

# noise2
ordershape_noise2 <- OrderShapeEM(pvals = p_value, order.var = noise2,
                                  OrderShapeEM.control(trace = TRUE))
ordershapeem_noise2 <- sapply(q_levels, function(q) {
  sum(ordershape_noise2$fdr <= q)
})

################## FDRreg #######################
set.seed(123)
fdr_theo_ae5 <- FDRreg(z_value, as.matrix(ae5), nulltype = 'theoretical')
rdrreg_ae5 <- sapply(q_levels, function(q) {
  length(which(fdr_theo_ae5$FDR <= q))
})

fdr_theo_pca5 <- FDRreg(z_value, as.matrix(pca5), nulltype = 'theoretical')
rdrreg_pca5 <- sapply(q_levels, function(q) {
  length(which(fdr_theo_pca5$FDR <= q))
})

fdr_theo_noise <- FDRreg(z_value, as.matrix(noise), nulltype = 'theoretical')
rdrreg_noise <- sapply(q_levels, function(q) {
  length(which(fdr_theo_noise$FDR <= q))
})

fdr_theo_noise2 <- FDRreg(z_value, as.matrix(noise2), nulltype = 'theoretical')
rdrreg_noise2 <- sapply(q_levels, function(q) {
  length(which(fdr_theo_noise2$FDR <= q))
})

################# IHW ####################
# Careful on IHW!!!!! May cause R terminated.
# autoencoder
ihw_x1_ae5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = ae5[,1], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x2_ae5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = ae5[,2], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x3_ae5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = ae5[,3], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x4_ae5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = ae5[,4], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x5_ae5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = ae5[,5], alpha = q, nbins = 5)
  rejections(ihw)
})

# pca
ihw_x1_pca5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = pca5[,1], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x2_pca5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = pca5[,2], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x3_pca5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = pca5[,3], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x4_pca5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = pca5[,4], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_x5_pca5 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = pca5[,5], alpha = q, nbins = 5)
  rejections(ihw)
})

# noise
ihw_noise <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = noise, alpha = q, nbins = 5)
  rejections(ihw)
})

# noise2
ihw_noise2 <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = noise2, alpha = q, nbins = 5)
  rejections(ihw)
})

##################### BH ######################
p_adjusted <- p.adjust(p_value, method = "BH")
bh_rejs <- sapply(q_levels, function(q) {
  length(which(p_adjusted <= q))
})

```

```{r, eval=FALSE}
############################# AE results only ###########################
# combine results into data frame
zhu_ae <- data.frame(
  q = q_levels,
  chai = chai_rejs_ae5,
  adapt_glm = adapt_glm_ns_ae5,
  adapt_gmm_p = adapt_gmm_p_ns_ae5,
  adapt_gmm_z = adapt_gmm_z_ns_ae5,
  OrderShapeEM_x1 = ordershape_ae5[, 1],
  OrderShapeEM_x2 = ordershape_ae5[, 2],
  OrderShapeEM_x3 = ordershape_ae5[, 3],
  OrderShapeEM_x4 = ordershape_ae5[, 4],
  OrderShapeEM_x5 = ordershape_ae5[, 5],
  FDRreg = rdrreg_ae5, 
  IHW_x1 = ihw_x1_ae5, 
  IHW_x2 = ihw_x2_ae5, 
  IHW_x3 = ihw_x3_ae5, 
  IHW_x4 = ihw_x4_ae5, 
  IHW_x5 = ihw_x5_ae5, 
  BH = bh_rejs
)

# Converge into long table for plot purpose
zhu_ae_long <- zhu_ae %>%
  pivot_longer(
    cols = -q,
    names_to = "Method",
    values_to = "Rejections"
  ) %>%
  mutate(Method = as.character(Method))

############################# AE and PCA results ###########################
zhu_ae_pca <- data.frame(
  q = q_levels,
  chai_ae = chai_rejs_ae5,
  chai_pca = chai_rejs_pca5,
  chai_noise1 = chai_rejs_noise,
  chai_noise2 = chai_rejs_noise2,
  adapt_glm_ae = adapt_glm_ns_ae5,
  adapt_glm_pca = adapt_glm_ns_pca5,
  adapt_glm_noise1 = adapt_glm_ns_noise,
  adapt_glm_noise2 = adapt_glm_ns_noise2,
  adapt_gmm_p_ae = adapt_gmm_p_ns_ae5,
  adapt_gmm_p_pca = adapt_gmm_p_ns_pca5,
  adapt_gmm_p_noise1 = adapt_gmm_p_ns_noise,
  adapt_gmm_p_noise2 = adapt_gmm_p_ns_noise2,
  adapt_gmm_z_ae = adapt_gmm_z_ns_ae5,
  adapt_gmm_z_pca = adapt_gmm_z_ns_pca5,
  adapt_gmm_z_noise1 = adapt_gmm_z_ns_noise,
  adapt_gmm_z_noise2 = adapt_gmm_z_ns_noise2,
  fdrreg_ae = rdrreg_ae5,
  fdrreg_pca = rdrreg_pca5,
  fdrreg_noise1 = rdrreg_noise,
  fdrreg_noise2 = rdrreg_noise2,
  bh = bh_rejs
)

# Converge into long table for plot purpose
zhu_ae_pca_long <- zhu_ae_pca %>%
  pivot_longer(
    cols = -q,
    names_to  = "name",
    values_to = "Rejections"
  ) %>%
  mutate(
    side = case_when(
      grepl("_ae$",   name) ~ "ae",
      grepl("_pca$",  name) ~ "pca",
      grepl("_noise1$", name) ~ "noise1",
      grepl("_noise2$", name) ~ "noise2",
      TRUE                    ~ "bh"        # BH has no suffix
    ),
    Method = case_when(
      side == "bh" ~ "bh",
      TRUE         ~ sub("_(ae|pca|noise1|noise2)$", "", name)
    )
  )
```

### Visualization

Define the plot function 
```{r}
# ---------- color helpers ----------
blend_color <- function(col, target = "white", amount = 0.4) {
  a <- grDevices::col2rgb(col) / 255
  b <- grDevices::col2rgb(target) / 255
  out <- (1 - amount) * a + amount * b
  grDevices::rgb(out[1,], out[2,], out[3,])
}

variant_shades <- function(base_col, K = 12) {
  grDevices::colorRampPalette(c(
    blend_color(base_col, "white", 0.55),
    base_col,
    blend_color(base_col, "black", 0.25)
  ))(K)
}

make_method_palette <- function(methods, K = 12) {
  fixed <- c(
    chai        = "red",
    adapt_glm   = "#FF7F0E",
    adapt_gmm_z = "#2CA02C",
    adapt_gmm_p = "#9467BD",
    FDRreg      = "#8C564B",
    BH          = "black"
  )

  family_base <- c(
    OrderShapeEM = "#1F77B4",  # blue family
    IHW          = "#17BECF"   # teal family
  )

  ord_shades <- variant_shades(family_base["OrderShapeEM"], K = K)
  ihw_shades <- variant_shades(family_base["IHW"],          K = K)

  methods <- unique(as.character(methods))
  pal <- setNames(rep("#7F7F7F", length(methods)), methods)  # fallback grey

  for (m in methods) {
    if (m %in% names(fixed)) {
      pal[m] <- fixed[m]
      next
    }

    base <- sub("_x\\d+$", "", m)
    idx  <- suppressWarnings(as.integer(sub(".*_x", "", m)))

    if (base %in% names(family_base)) {
      if (is.na(idx)) {
        pal[m] <- family_base[base]  # e.g. "OrderShapeEM" with no suffix
      } else {
        idx <- max(1L, min(idx, K))
        pal[m] <- if (base == "OrderShapeEM") ord_shades[idx] else ihw_shades[idx]
      }
    }
  }

  pal
}

method_order <- function(methods) {
  methods <- unique(as.character(methods))
  core <- c("chai","adapt_glm","adapt_gmm_z","adapt_gmm_p","FDRreg")

  sort_family <- function(base) {
    fam <- methods[grepl(paste0("^", base, "(_x\\d+)?$"), methods)]
    base_present <- base %in% fam
    vars <- fam[grepl("_x\\d+$", fam)]
    if (length(vars)) {
      idx <- suppressWarnings(as.integer(sub(".*_x", "", vars)))
      vars <- vars[order(idx)]
    }
    c(if (base_present) base else character(0), vars)
  }

  out <- c(core, sort_family("OrderShapeEM"), sort_family("IHW"), "BH")
  out <- unique(out)
  out[out %in% methods]
}

# ---------- plotting function (labels, no legend) ----------
library(ggrepel)
library(stringr)

plot_rejections_vs_q <- function(long_table, # title = "Number of Discoveries vs q",
                                 K_variants = 12, x_breaks = NULL, method_font = 5.5) {
  
  ord <- method_order(long_table$Method)
  long_table <- long_table %>% mutate(Method = factor(Method, levels = ord))
  pal <- make_method_palette(levels(long_table$Method), K = K_variants)
  
  endpoints <- long_table %>%
    group_by(Method) %>%
    filter(q == max(q, na.rm = TRUE)) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    mutate(label = as.character(Method))
  
  x_rng   <- range(long_table$q, na.rm = TRUE)
  x_nudge <- diff(x_rng) * 0.03
  
  p <- ggplot(long_table, aes(x = q, y = Rejections, group = Method, color = Method)) +
    geom_line(linewidth = 0.9, na.rm = TRUE) +
    geom_point(size = 1.8, na.rm = TRUE) +
    ggrepel::geom_text_repel(
      data = endpoints,
      aes(label = label),
      size = method_font,
      nudge_x = x_nudge,
      hjust = 0,
      direction = "y",
      segment.size = 0.2,
      box.padding = 0.15,
      point.padding = 0.10,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 1,
      show.legend = FALSE
    ) +
    scale_color_manual(values = pal, breaks = ord) +
    labs( #title = title, 
      x = "q level", y = "Number of Rejections") +
    coord_cartesian(clip = "off") +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.margin = margin(5.5, 60, 5.5, 5.5),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      axis.text.x  = element_text(size = 12),
      axis.text.y  = element_text(size = 12)
    )
  
  if (!is.null(x_breaks)) {
    p <- p + scale_x_continuous(breaks = x_breaks, expand = expansion(mult = c(0.02, 0.18)))
  } else {
    p <- p + scale_x_continuous(expand = expansion(mult = c(0.02, 0.18)))
  }
  
  p
}
```

Line chart of number or selections by every method at different FDR q level. 

```{r}
p_zhu_ae <- plot_rejections_vs_q(zhu_ae_long,
                              x_breaks = sort(unique(zhu_ae_long$q)))
```

```{r, fig.width=8, fig.height=5, out.width='60%', warning=FALSE}
p_zhu_ae
```


```{r}
zhu_ae_pca_long$side <- factor(zhu_ae_pca_long$side, levels = c("ae", "pca", "noise1", "noise2"))

# Only methods with the three side-info types (drop BH panel)
zhu_long_methods <- zhu_ae_pca_long %>%
  filter(side != "bh", Method != "bh")

# --- Put 'chai' as the first facet ----------------------------------------
method_levels <- c("chai", setdiff(unique(as.character(zhu_long_methods$Method)), "chai"))

zhu_long_methods <- zhu_long_methods %>%
  mutate(Method = factor(Method, levels = method_levels))

# --- End-of-line label positions (one per Method C side) -------------------
endpoints <- zhu_long_methods %>%
  arrange(Method, side, q) %>%
  group_by(Method, side) %>%
  slice_tail(n = 1) %>%
  ungroup()

dx <- diff(range(zhu_long_methods$q))

# BH baseline (as before)
bh_line <- zhu_ae_pca_long %>%
  dplyr::filter(Method == "bh") %>%
  dplyr::select(q, Rejections) %>%
  tidyr::crossing(Method = method_levels) %>%
  # IMPORTANT: make Method a factor with same levels
  mutate(Method = factor(Method, levels = method_levels))

# Endpoints for BH (one per Method)
bh_endpoints <- bh_line %>%
  group_by(Method) %>%
  slice_max(q, with_ties = FALSE) %>%
  ungroup()

p_ae_pca <- ggplot(zhu_long_methods,
                   aes(x = q, y = Rejections, color = side)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text_repel(
    data = endpoints,
    aes(label = side),
    nudge_x      = dx * 0.03,
    hjust        = 0,
    direction    = "y",
    segment.size = 0.2,
    box.padding  = 0.1,
    point.padding= 0.1,
    show.legend  = FALSE
  ) +
  # BH dashed line
  geom_line(
    data        = bh_line,
    aes(x = q, y = Rejections, group = 1),
    inherit.aes = FALSE,
    color       = "black",
    linewidth   = 0.8,
    linetype    = "dashed"
  ) +
  # "bh" label
  geom_text(
    data        = bh_endpoints,
    aes(x = q, y = Rejections, label = "bh"),
    inherit.aes = FALSE,
    nudge_x     = dx * 0.03,
    hjust       = 0,
    vjust       = 0.5,
    size        = 3,
    color       = "black"
  ) +
  facet_wrap(~ Method) +  # <-- remove scales = "free_y"
  scale_x_continuous(breaks = unique(zhu_long_methods$q)) +
  scale_y_continuous(limits = c(0, max(zhu_long_methods$Rejections))) +
  scale_color_manual(
    values = c(
      ae    = "red",
      pca   = "blue",
      noise1 = "grey50",
      noise2 = "grey70"
    )
  ) +
  labs(
    # title  = "Number of Discoveries at different q level",
    x      = "q level",
    y      = "Number of Rejections",
    color  = "Side information: "
  ) +
  coord_cartesian(clip = "off") +
  expand_limits(x = max(zhu_long_methods$q) + dx * 0.05) +
  theme_minimal() +
  theme(
    strip.text      = element_text(face = "bold"),
    legend.position = "bottom",
    plot.margin     = margin(5.5, 60, 5.5, 5.5)
  )
```

```{r, fig.width=8, fig.height=5, out.width='100%', warning=FALSE}
p_ae_pca
```


```{r, include=FALSE, eval=FALSE}
# save(p_value, z_value, outcome, t_abun, 
#      ae5, pca5, noise, noise2,
#      chai_zhu_ae5, chai_zhu_pca5, chai_zhu_noise, chai_zhu_noise2, 
#      zhu_ae, zhu_ae_long,
#      zhu_ae_pca, zhu_ae_pca_long,
#      file = "./chai_env/Schizophrenia_env.RData")

# ggsave("./plots/Zhu/Zhu_AE_updated.png", plot = p_zhu_ae, width = 8, height = 5, dpi = 300)

# pdf("./plots/Zhu/Zhu_AE_updated.pdf", width = 8, height = 5)
# plot(p_zhu_ae)
# dev.off()


# ggsave("./plots/Zhu/Zhu_AE5_vs_PCA5_vs_noise.png", plot = p_ae_pca, width = 8, height = 5, dpi = 300)

# pdf("./plots/Zhu/Zhu_AE5_vs_PCA5_vs_noise.pdf", width = 8, height = 5)
# plot(p_ae_pca)
# dev.off()
```
