---
title: "Melanoma anti-PD-1 Real Data Analysis - PCoA"
author: "Ziyi Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

## Overview

This document reproduces the Melanoma anti-PD-1 real data analysis using
PCoA of phylogentic tree structure as side covariates (auxiliary information).

**Key settings (matching the manuscript figure):**

| Parameter | Value |
|-----------|-------|
| Dataset | Gopalakrishnan et al. (Melanoma, 16S rRNA) |
| Target taxonomic rank | Family |
| Primary test statistic | Wilcoxon z-statistic / DESeq2 |
| FDR methods compared | chai, AdaPT-GLM, AdaPT-GMM, OrderShapeEM, FDRreg, IHW, BH |


**Working directory:** Knit this file from (`chai_project/scripts/`).

---

## 1. Install Dependencies

Run once in a fresh R session before knitting:

```{r install, eval=FALSE}
# CRAN + Bioconductor + GitHub dependencies
source("install_packages.R")

# chai package (not on CRAN)
remotes::install_github("ziyiwang726/chai")
```

---

## 2. Load Packages

```{r packages, include=TRUE, warning=FALSE}
suppressPackageStartupMessages({
  library(chai)
  library(knitr)
  library(CATMicrobiome)
  library(mixtools)
  library(KScorrect)
  library(adaptMT)
  library(AdaPTGMM)
  library(splines2)
  library(dplyr)
  library(ape)
  library(phyloseq)
  library(ggplot2)
  library(tidyr)
  library(coin)
  library(ggrepel)
  library(IHW)
  library(FDRreg)
  library(OrderShapeEM)
  library(DESeq2)
  library(ggtree)
  library(ggtreeExtra)
  library(ggnewscale)
})

```

---

## 3. Load the data and pre-processing the data

**Note**: 

The pro-processing part and model fitting chunk could be time-consuming. We saved the processed data together with the organized results after fitting the models. When knitting this HTML file, we loaded the **Gopalakrishnan_16s_PCoA_env.RData** file directly in sectcion 4 and generated the plots from it. 

However, for verification or reproducibility, you can also run each pre-processing chunk and/or model-fitting chunk one by one, while skipping the chunk that loads the .RData file. This should produce the same results.

```{r, eval=FALSE}
#################### Load data from package CATMicrobiome ##############
otuPath <- system.file("extdata","d1OTUtable.csv", 
                       package = "CATMicrobiome")
otutable <- read.csv(otuPath,header=TRUE,row.names = 1)

taxonomyPath <- system.file("extdata","d1Taxonomy.csv", 
                            package = "CATMicrobiome")
taxonomy <- read.csv(taxonomyPath,header=TRUE,row.names = 1)

metaPath <- system.file("extdata","d1Meta.csv", 
                        package = "CATMicrobiome")
metaData <- read.csv(metaPath,header=TRUE,row.names = 1)

treePath <- system.file("extdata","d1Tree.tree", 
                        package = "CATMicrobiome")
tree <- read.tree(treePath)

otus_tree   <- tree$tip.label       # 1437
otus_tax    <- rownames(taxonomy)   # 1455
otus_keep   <- intersect(otus_tree, otus_tax)   # Keep 1437

# Prune tree and taxonomy
tree <- keep.tip(tree, otus_keep)
taxonomy <- taxonomy[otus_keep, , drop = FALSE]
taxonomy <- taxonomy[match(tree$tip.label, rownames(taxonomy)), , drop = FALSE]  # reorder to tree tips
otutable <- otutable[otus_keep, , drop = FALSE]
# 1437 tips

OTU <- otu_table(as.matrix(otutable), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(taxonomy))
SAM <- sample_data(metaData)
TR  <- phy_tree(tree)

############### Build phyloseq-class objects ################
ps <- phyloseq(OTU, TAX, SAM, TR)

###################### Aggregate into Family level #######################
family <- tax_glom(ps, taxrank='family')

###################### Abundance table #######################
cs <- colSums(otu_table(family))
cs[cs == 0] <- 1
family_abun <- sweep(otu_table(family), 2, cs, "/") * 100    # family x subjects

# Transpose for modeling (subjects × species)
t_abun <- t(family_abun)     # 43 x 75

########################## Outcome ###################################
outcome <- factor(sample_data(family)$BinOutcomes, levels = c("NR","R"))
table(outcome)
```

---

## 4. PCoA Auxiliary Covariates

Generating the side information - Calculate the cophenetic distances from phylogentic tree at family level, and then fit into PCoA.


```{r, eval=FALSE}
# Calculate the cophenetic distances from phylogentic tree at family level
family_D <- cophenetic(phy_tree(family))
# 76 x 76

# Fit the PCoA 
family_pcoa <- cmdscale(family_D, k = 2)
```

---

## 5. Wilcoxon Test Statistics

```{r, eval=FALSE}
p_z_values <- sapply(colnames(t_abun), function(feature) {
  features <- t_abun[, feature, drop = FALSE]
  wilcox <- wilcox_test(features[,1] ~ outcome)

  p_value <- pvalue(wilcox)
  z_value <- statistic(wilcox)

  return(c(p_value = p_value, z_value = z_value))
}, simplify = "matrix")
p_z_values <- t(p_z_values)

p_value <- p_z_values[,1]
z_value <- p_z_values[,2]

```


---

## 6. Load the processed data

The processed data included: 

- **z_value**: the z-statistics from wilcoxon test of each family between NR vs R.
- **p_value**: the p-value from wilcoxon test of each family between NR vs R.
- **family_pcoa**: the side information. The two dimensional PCoA of cophenetic distances from phylogentic tree at family level.
- **t_abun**: the relative abundance table of this dataset at family level.
- **outcome**: the outcome (NR or R) of this dataset. 
- **family**: the phyloseq-class objects saving otu table, taxon table, metadata information and tree structure.

- **chai_family_pcoa**: \code(chai) output with wilcoxon z_statistics and family_pcoa (side information) as inputs.
- **gopa_fam_16s**: the summary table saving the number of selections of every method at different q, with wilcoxon z + family_pcoa.
- **gopa_fam_16s_long**: the long table version of gopa_fam_16s for plot purpose.

- **DES_stat**: the DESeq2 wald statistics of each family between NR vs R.
- **DES_p**: the DESeq2 p-value of each family between NR vs R.
- **res**: the DESeq2 analysis results at q = 0.05 target FDR level.

- **chai_family_pcoa_DES**: \code(chai) output with DESeq2 wald_statistics and family_pcoa (side information) as inputs.
- **gopa_fam_16s_DES**: the summary table saving the number of selections of every method at different q, with DESeq2 wald-stat + family_pcoa.
- **gopa_fam_16s_DES_long**: the long table version of gopa_fam_16s_DES for plotting purpose.


```{r}
load("../data/chai_env/Gopalakrishnan_16s_PCoA_env.RData")
```

---

## 7. Multiple Testing Methods

### 7a. BH Fit the models

```{r bh, eval=FALSE, warning=FALSE}
q_levels <- seq(0.01, 0.10, by = 0.01)

########################## BH ##########################
p_adjusted <- p.adjust(p_value, method = "BH")
bh_rejs <- sapply(q_levels, function(q) {
  length(which(p_adjusted <= q))
})
```


### 7b. chai

```{r chai, eval=FALSE, warning=FALSE}
set.seed(123)
chai_family_pcoa <- chai(z_value, family_pcoa, B = 1000)
chai_rejs <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_family_pcoa$clFDR, q, 1))
})
```

### 7c. AdaPT-GLM

```{r adapt-glm, eval=FALSE}
x <- data.frame(family_pcoa)
formula_ns <- list("~ns(X1, df = 2) + ns(X2, df = 2)",
                   "~ns(X1, df = 4) + ns(X2, df = 4)",
                   "~ns(X1, df = 6) + ns(X2, df = 6)")

glm_ns <- adapt_glm(x, pvals = p_value, alphas=q_levels, 
                    pi_formulas = formula_ns, mu_formulas = formula_ns)
adapt_glm_ns_rej <- glm_ns$nrejs[1:10]
```

### 7d. AdaPT-GMM (p-values)

```{r adapt-gmm-p, eval=FALSE}
gmm_ns_p <- adapt_gmm(x, pvals = p_value,  alphas=q_levels,
                      beta_formulas = formula_ns)
adapt_gmm_p_ns_rej <- gmm_ns_p$nrejs[1:10]
```

### 7e. AdaPT-GMM (z-statistics)

```{r adapt-gmm-z, eval=FALSE}
gmm_ns_z <- adapt_gmm(x, z = z_value, alphas=q_levels,
                      beta_formulas = formula_ns, testing = "two_sided")
adapt_gmm_z_ns_rej <- gmm_ns_z$nrejs[1:10]
```

### 7f. OrderShapeEM

```{r ordershape, eval=FALSE}
ordershape_x1 <- OrderShapeEM(pvals = p_value, order.var = family_pcoa[,1], 
                              OrderShapeEM.control(trace = TRUE))
ordershapeem_x1_rejs <- sapply(q_levels, function(q) {
  sum(ordershape_x1$fdr <= q)
})

ordershape_x2 <- OrderShapeEM(pvals = p_value, order.var = family_pcoa[,2], 
                              OrderShapeEM.control(trace = TRUE))
ordershapeem_x2_rejs <- sapply(q_levels, function(q) {
  sum(ordershape_x2$fdr <= q)
})
```

### 7g. FDRreg

```{r fdrreg, eval=FALSE}
set.seed(123)
fdr_theo <- FDRreg(z_value, family_pcoa, nulltype = 'theoretical')
rdrreg_rejs <- sapply(q_levels, function(q) {
  length(which(fdr_theo$FDR <= q))
})
```

### 7h. IHW
** Note: ** Careful on running IHW.

```{r ihw, eval=FALSE, message=FALSE}
set.seed(123)
ihw_1_rejs <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = family_pcoa[,1], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_2_rejs <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = p_value, covariates = family_pcoa[,2], alpha = q, nbins = 5)
  rejections(ihw)
})
```

---

## 8. Main Figure — Number of Discoveries vs FDR Threshold

```{r, eval=FALSE, fig.width=10, fig.height=5}
gopa_fam_16s <- data.frame(
  q = q_levels,
  chai = chai_rejs,
  adapt_glm = adapt_glm_ns_rej,
  adapt_gmm_p = adapt_gmm_p_ns_rej,
  adapt_gmm_z = adapt_gmm_z_ns_rej,
  OrderShapeEM_x1 = ordershapeem_x1_rejs,
  OrderShapeEM_x2 = ordershapeem_x2_rejs,
  FDRreg = rdrreg_rejs,
  IHW_x1 = ihw_1_rejs,
  IHW_x2 = ihw_2_rejs,
  BH = bh_rejs
)


# converge into long format for visualization
gopa_fam_16s_long <- gopa_fam_16s %>%
  tidyr::pivot_longer(
    cols = -q,
    names_to = "Method",
    values_to = "Rejections"
  ) %>%
  mutate(Method = as.character(Method))
```

Line chart of number or selections by every method at different FDR q level. 

```{r fig1}
source("consistant_color.R")

p_PCoA <- plot_rejections_vs_q(gopa_fam_16s_long,
                              x_breaks = sort(unique(gopa_fam_16s_long$q)))

```

```{r, fig.width=8, fig.height=5, out.width='100%', warning=FALSE}
p_PCoA
```


---

## 9. DESeq2 Statistics

```{r, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
cts <- as(otu_table(family), "matrix")

md  <- as(sample_data(family), "data.frame")
md$BinOutcomes <- factor(md$BinOutcomes, levels = c("NR", "R"))

dds <- DESeqDataSetFromMatrix(countData = cts, colData = md,
                              design = ~ BinOutcomes)
dds_res <- DESeq(dds)
res <- results(dds_res, alpha = 0.05)
summary(res)

DES_stat <- res$stat
DES_p <- res$pvalue

```

---

## 10. Multiple Testing Methods
### 10a. BH Fit the models

```{r bh_des, eval=FALSE, warning=FALSE}
q_levels <- seq(0.01, 0.10, by = 0.01)

########################## BH ##########################
p_adjusted_DES <- p.adjust(DES_p, method = "BH")
bh_rejs_DES <- sapply(q_levels, function(q) {
  length(which(p_adjusted_DES <= q))
})
```

### 10b. chai

```{r chai_des, eval=FALSE, warning=FALSE}
set.seed(123)
chai_family_pcoa_DES <- chai(DES_stat, family_pcoa, B = 1000)
chai_rejs_DES <- sapply(q_levels, function(q) {
  length(clfdrselect(chai_family_pcoa_DES$clFDR, q, 1))
})
```

### 10c. AdaPT-GLM

```{r adapt-glm_des, eval=FALSE}
x <- data.frame(family_pcoa)
formula_ns <- list("~ns(X1, df = 2) + ns(X2, df = 2)",
                   "~ns(X1, df = 4) + ns(X2, df = 4)",
                   "~ns(X1, df = 6) + ns(X2, df = 6)")

glm_ns_DES <- adapt_glm(x, pvals = DES_p, alphas=q_levels,
                        pi_formulas = formula_ns, mu_formulas = formula_ns)

adapt_glm_ns_rej_DES <- glm_ns_DES$nrejs[1:10]
```

### 10d. AdaPT-GMM (p-values)

```{r adapt-gmm-p_des, eval=FALSE}
gmm_ns_p_DES <- adapt_gmm(x, pvals = DES_p,  alphas=q_levels,
                      beta_formulas = formula_ns)
adapt_gmm_p_ns_rej_DES <- gmm_ns_p_DES$nrejs[1:10]
```

### 10e. AdaPT-GMM (z-statistics)

```{r adapt-gmm-z_des, eval=FALSE}
gmm_ns_z_DES <- adapt_gmm(x, z = DES_stat, alphas=q_levels,
                      beta_formulas = formula_ns, testing = "two_sided")
adapt_gmm_z_ns_rej_DES <- gmm_ns_z_DES$nrejs[1:10]
```

### 10f. OrderShapeEM

```{r ordershape_des, eval=FALSE}
ordershape_x1_DES <- OrderShapeEM(pvals = DES_p, order.var = family_pcoa[,1], 
                              OrderShapeEM.control(trace = TRUE))
ordershapeem_x1_rejs_DES <- sapply(q_levels, function(q) {
  sum(ordershape_x1_DES$fdr <= q)
})

ordershape_x2_DES <- OrderShapeEM(pvals = DES_p, order.var = family_pcoa[,2], 
                              OrderShapeEM.control(trace = TRUE))
ordershapeem_x2_rejs_DES <- sapply(q_levels, function(q) {
  sum(ordershape_x2_DES$fdr <= q)
})
```

### 10g. FDRreg

```{r fdrreg_des, eval=FALSE}
set.seed(123)
fdr_theo_DES <- FDRreg(DES_stat, family_pcoa, nulltype = 'theoretical')
rdrreg_rejs_DES <- sapply(q_levels, function(q) {
  length(which(fdr_theo_DES$FDR <= q))
})
```

### 10h. IHW
** Note: ** Careful on running IHW.

```{r ihw_des, eval=FALSE, message=FALSE}
set.seed(123)
ihw_1_rejs_DES <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = DES_p, covariates = family_pcoa[,1], alpha = q, nbins = 5)
  rejections(ihw)
})

ihw_2_rejs_DES <- sapply(q_levels, function(q) {
  ihw <- ihw(pvalues = DES_p, covariates = family_pcoa[,2], alpha = q, nbins = 5)
  rejections(ihw)
})
```


---

## 11. Main Figure

```{r, eval=FALSE, fig.width=10, fig.height=5}
gopa_fam_16s_DES <- data.frame(
  q = q_levels,
  chai = chai_rejs_DES,
  adapt_glm = adapt_glm_ns_rej_DES,
  adapt_gmm_p = adapt_gmm_p_ns_rej_DES,
  adapt_gmm_z = adapt_gmm_z_ns_rej_DES,
  OrderShapeEM_x1 = ordershapeem_x1_rejs_DES,
  OrderShapeEM_x2 = ordershapeem_x2_rejs_DES,
  FDRreg = rdrreg_rejs_DES,
  IHW_x1 = ihw_1_rejs_DES,
  IHW_x2 = ihw_2_rejs_DES,
  BH = bh_rejs_DES
)


# into long format
gopa_fam_16s_DES_long <- gopa_fam_16s_DES %>%
  tidyr::pivot_longer(
    cols = -q,
    names_to = "Method",
    values_to = "Rejections"
  ) %>%
  mutate(Method = as.character(Method))
```

### 11a. Number of Discoveries vs FDR Threshold

```{r fig2}
p_PCoA_DES <- plot_rejections_vs_q(gopa_fam_16s_DES_long,
                                   x_breaks = sort(unique(gopa_fam_16s_DES_long$q)))

```

```{r, fig.width=8, fig.height=5, out.width='100%', warning=FALSE}
p_PCoA_DES
```


### 11b. The phylogentic tree

```{r, message=FALSE, warning=FALSE}
# Define a function to clean the feature name
clean_taxon <- function(x) {
  x %>%
    as.character() %>%
    str_replace("^.*?__", "") %>%        # Remove k__/f__/g__
    str_replace_all("\\[|\\]", "") %>%   # Remove []
    str_squish()                         # Remove extra space
}

# Get the indices of the features that selected by chai, at q = 0.05
chai_DES_sel <- clfdrselect(chai_family_pcoa_DES$clFDR, 0.05, 1)
# Get the chai selected OTU IDs
sel_otus_DES <- rownames(res)[chai_DES_sel] 

# Get the indices of the features that selected by BH, at q = 0.05
p_adjusted_DES <- p.adjust(DES_p, method = "BH")
DES_bh <- which(p_adjusted_DES <= 0.05)
# Get the BH selected OTU IDs
sel_otus_BH <- rownames(res)[DES_bh]

# map labels for ALL tips (not only selected)
tax_all <- as.data.frame(tax_table(family))
tax_all$OTU <- rownames(tax_all)

# Prepare the annotation
rank_to_show <- "family"
anno_DES <- tax_all %>%
  mutate(
    # clean the family names
    raw = .data[[rank_to_show]],
    clean = clean_taxon(raw),
    tip_label = ifelse(is.na(clean) | clean == "", OTU, clean),

    # clean the phylum names (for annotation)
    phylum_clean = clean_taxon(phylum),

    # tip color
    tip_color = dplyr::case_when(
      OTU %in% sel_otus_DES & OTU %in% sel_otus_BH ~ "both",
      OTU %in% sel_otus_DES & !(OTU %in% sel_otus_BH) ~ "chai_only",
      !(OTU %in% sel_otus_DES) & OTU %in% sel_otus_BH ~ "BH_only",
      TRUE ~ "neither"
    )  ) %>%
  dplyr::select(OTU, tip_label, phylum_clean, tip_color) %>%
  distinct(OTU, .keep_all = TRUE)

tr_all <- phy_tree(family)
tr <- ape::as.phylo(tr_all)

# Get the basic circular tree plot
p <- ggtree::ggtree(tr, layout = "circular", size = 0.5)

p1 <- p %<+% anno_DES

# Add phylum annotation
p2 <- p1 +
  new_scale_fill() +
  ggtreeExtra::geom_fruit(
    geom = geom_tile,
    mapping = aes(fill = phylum_clean), 
    width = 0.2,
    offset = 0.05
  ) +
  labs(fill = "phylum")

# Add tip labels
p3 <- p2 +
  geom_tiplab(
    aes(label = tip_label, color = tip_color),
    size = 2,
    offset = 0.02,          # distance between text and color block
    show.legend = FALSE
  ) +
  scale_color_manual(
    name   = "Tip label group",
    values = c(chai_only="#E69F00", BH_only="#0072B2", both="#CC79A7", neither="#666666"),
    breaks = c("chai_only", "BH_only", "both", "neither"),
    labels = c("Only in chai", "Only in DESeq2", "In both", "Neither"),
    drop   = FALSE
  ) +
  # labs(title = "Phylogenetic tree with selected families by chai on DESeq2 z with PCoA") +
  theme(legend.position = "right")

legend_df <- data.frame(
  x = NA_real_, y = NA_real_,
  tip_color = factor(c("chai_only","BH_only","both","neither"),
                     levels = c("chai_only","BH_only","both","neither"))
)

# Add tip label group legend
p4 <- p3 +
  geom_point(
    data = legend_df,
    aes(x = x, y = y, color = tip_color),
    inherit.aes = FALSE,
    size = 3
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1))) +
  theme(legend.position = "right")
```

```{r, fig.width=8, fig.height=8, out.width='100%', warning=FALSE}
p4
```


---

## Session Info

```{r session-info}
sessionInfo()
```
