---
title: "Melanoma Real Data Analysis — LLM Side Information"
author: "Yushu Shi, Ziyi Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

## Overview

This document reproduces the Melanoma anti-PD-1 real data analysis using
LLM-derived literature evidence as side covariates (auxiliary information).

**Key settings (matching the manuscript figure):**

| Parameter | Value |
|-----------|-------|
| Dataset | Gopalakrishnan et al. (Melanoma, 16S rRNA) |
| Target taxonomic rank | Family |
| LLM auxiliary source | Family and below (combined provider) |
| Aggregation mode | **Weighted** Stouffer z |
| Primary test statistic | Wilcoxon z-statistic |
| FDR methods compared | chai, AdaPT-GLM, AdaPT-GMM, OrderShapeEM, FDRreg, IHW, BH |

**Input LLM auxiliary file:**
```
chai_project/Gopalakrishnan/sideCov/combined/weighted/selectionTargetFamily/
  d1Taxonomy_family_ABCD_family_and_below_weighted.csv
```
This file is pre-computed by the LLM pipeline (`chai_project/Gopalakrishnan/LLMCode/`)
and is already included in the repository. You do **not** need to re-run
the LLM pipeline to reproduce these results.

**Working directory:** Knit this file from (`chai_project/scripts/`).

---

## 1. Install Dependencies

Run once in a fresh R session before knitting:

```{r install, eval=FALSE}
# CRAN + Bioconductor + GitHub dependencies
source("Gopalakrishnan/install_packages.R")

# chai package (not on CRAN)
remotes::install_github("ziyiwang726/chai")
```

---

## 2. Load Packages

```{r packages}
library(chai)
library(ape)
library(CATMicrobiome)
library(coin)
library(phyloseq)
library(adaptMT)
library(AdaPTGMM)
library(splines)
library(splines2)
library(IHW)
library(FDRreg)
library(OrderShapeEM)
library(auctestr)
library(dplyr)
library(tibble)
library(stringr)
library(tidyr)
library(readr)
library(ggplot2)
library(ggrepel)

```

---

## 3. Configuration

```{r config}
# ── Analysis parameters (matching manuscript figure) ──────────────────────────
TARGET_RANK <- "family"
AUX_SOURCE  <- "family"
AUX_MODE    <- "weighted"
PROVIDER    <- "combined"

# ── Input: LLM auxiliary covariate CSV ────────────────────────────────────────
aux_csv <- file.path(
  "../Gopalakrishnan", "sideCov", "combined", "weighted",
  "selectionTargetFamily",
  "d1Taxonomy_family_ABCD_family_and_below_weighted.csv"
)

stopifnot("LLM auxiliary CSV not found — check working directory" = file.exists(aux_csv))

# ── Reproducibility ───────────────────────────────────────────────────────────
SEED     <- 123L
B_CHAI   <- 1000L
Q_LEVELS <- seq(0.01, 0.10, by = 0.01)

set.seed(SEED)
```

---

## 4. Helper Functions

```{r helpers}
clean_taxon <- function(x) {
  x %>%
    as.character() %>%
    str_replace("^.*?__", "") %>%
    str_replace_all("\\[|\\]", "") %>%
    str_squish()
}

compute_pz_wilcox <- function(t_abun, outcome) {
  stopifnot(nrow(t_abun) == length(outcome))
  res <- sapply(colnames(t_abun), function(feat) {
    wt <- wilcox_test(t_abun[, feat] ~ outcome)
    c(p_value = as.numeric(pvalue(wt)), z_value = as.numeric(statistic(wt)))
  }, simplify = "matrix")
  out <- as.data.frame(t(res))
  out$feature <- rownames(out)
  rownames(out) <- NULL
  out
}

make_main_df <- function(t_abun, pz_df, ps_obj, rank) {
  tax_df   <- as.data.frame(tax_table(ps_obj)) %>% rownames_to_column("otu_rep")
  rank_col <- names(tax_df)[tolower(names(tax_df)) == tolower(rank)]
  if (length(rank_col) != 1)
    stop("Cannot find unique taxonomy column for rank = '", rank, "'")
  tax_map <- tax_df %>%
    transmute(otu_rep, rank_value = as.character(.data[[rank_col]]))
  pz_map <- pz_df %>%
    transmute(otu_rep = feature,
              p_main  = as.numeric(p_value),
              z_main  = as.numeric(z_value))
  pz_map %>%
    left_join(tax_map, by = "otu_rep") %>%
    mutate(taxon_key = clean_taxon(rank_value))
}

stouffer_nonzero <- function(v) {
  v2 <- v[!is.na(v) & v != 0]
  if (!length(v2)) return(0)
  auctestr::stouffer_z(v2, ignore.na = FALSE)
}

make_aux_tax_level <- function(aux_otu, level, aux_cols = c("A", "B", "C", "D")) {
  present_cols <- intersect(aux_cols, names(aux_otu))
  level_col    <- names(aux_otu)[tolower(names(aux_otu)) == tolower(level)]
  if (length(level_col) != 1)
    stop("Cannot find level column '", level, "' in auxiliary CSV")
  aux_otu %>%
    filter(!is.na(.data[[level_col]]), .data[[level_col]] != "") %>%
    group_by(.data[[level_col]]) %>%
    summarise(across(all_of(present_cols), stouffer_nonzero, .names = "x{col}"),
              .groups = "drop") %>%
    mutate(taxon_key = clean_taxon(.data[[level_col]]))
}

build_ns_formulas <- function(X, dfs = c(2, 4, 6)) {
  n_unique <- vapply(X, function(col) length(unique(col[is.finite(col)])), integer(1))
  varying  <- names(n_unique)[n_unique > 1]
  lapply(dfs, function(df_req) {
    if (!length(varying)) return(~ 1)
    terms <- vapply(varying, function(v) {
      nu <- n_unique[[v]]
      if (nu <= 2) v else sprintf("ns(%s, df=%d)", v, min(df_req, nu - 1))
    }, character(1))
    as.formula(paste("~", paste(terms, collapse = " + ")))
  })
}

compute_ihw_rejs <- function(p, covariate, name, q_levels) {
  if (length(unique(covariate[is.finite(covariate)])) <= 1) {
    message("Skipping IHW for ", name, " (constant covariate)")
    return(rep(NA_integer_, length(q_levels)))
  }
  tryCatch(
    sapply(q_levels, function(q)
      rejections(ihw(pvalues = p, covariates = covariate, alpha = q, nbins = 5))),
    error = function(e) {
      warning("IHW failed for ", name, ": ", e$message)
      rep(NA_integer_, length(q_levels))
    }
  )
}
```

---

## 5. Load and Pre-process Data

```{r data}
otutable <- read.csv(
  system.file("extdata", "d1OTUtable.csv", package = "CATMicrobiome"),
  header = TRUE, row.names = 1)
taxonomy <- read.csv(
  system.file("extdata", "d1Taxonomy.csv", package = "CATMicrobiome"),
  header = TRUE, row.names = 1)
metaData <- read.csv(
  system.file("extdata", "d1Meta.csv", package = "CATMicrobiome"),
  header = TRUE, row.names = 1)
tree <- read.tree(
  system.file("extdata", "d1Tree.tree", package = "CATMicrobiome"))

otus_keep <- intersect(tree$tip.label, rownames(taxonomy))
tree      <- keep.tip(tree, otus_keep)
taxonomy  <- taxonomy[match(tree$tip.label, rownames(taxonomy)), , drop = FALSE]
otutable  <- otutable[otus_keep, , drop = FALSE]

ps <- phyloseq(
  otu_table(as.matrix(otutable), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxonomy)),
  sample_data(metaData),
  phy_tree(tree)
)
```

```{r agglomerate}
family_ps     <- tax_glom(ps, taxrank = "family")

cs            <- colSums(otu_table(family_ps)); cs[cs == 0] <- 1
family_abun   <- sweep(otu_table(family_ps), 2, cs, "/") * 100
t_family_abun <- t(family_abun)

outcome <- factor(sample_data(family_ps)$BinOutcomes, levels = c("NR", "R"))
table(outcome)
```

---

## 6. Wilcoxon Test Statistics

```{r wilcoxon}
family_pz     <- compute_pz_wilcox(t_family_abun, outcome)
family_pz_tax <- make_main_df(t_family_abun, family_pz, family_ps, rank = "family")
```

---

## 7. LLM Auxiliary Covariates

```{r aux}
family_aux_otu <- read_csv(aux_csv, show_col_types = FALSE)

fam_aux <- make_aux_tax_level(family_aux_otu, level = "family")

model_covariates <- c("xA", "xB", "xC")
df_model <- family_pz_tax %>%
  dplyr::left_join(fam_aux %>% dplyr::select(taxon_key, all_of(model_covariates)),
            by = "taxon_key") %>%
  dplyr::mutate(across(all_of(model_covariates), ~ replace_na(.x, 0)))

keep     <- complete.cases(df_model[, c("z_main", "p_main", model_covariates)])
df_model <- df_model[keep, ]

z <- df_model$z_main
p <- df_model$p_main
X <- as.data.frame(df_model[, model_covariates])
```

---

## 8. Multiple Testing Methods

### 8a. BH

```{r bh}
p_adjusted <- p.adjust(p, method = "BH")
bh_rejs    <- sapply(Q_LEVELS, function(q) sum(p_adjusted <= q))
```

### 8b. chai

```{r chai}
set.seed(SEED)
chai_fit  <- chai(z, X, K_vec = 2:10, B = B_CHAI)
chai_lfdr <- if (!is.null(chai_fit$lFDR)) chai_fit$lFDR else chai_fit$clFDR
chai_rejs <- sapply(Q_LEVELS, function(q) length(clfdrselect(chai_lfdr, q, 1)))
```

### 8c. AdaPT-GLM

```{r adapt-glm}
formula_ns     <- build_ns_formulas(X)
glm_ns         <- adapt_glm(X, pvals = p, alphas = Q_LEVELS,
                             pi_formulas = formula_ns, mu_formulas = formula_ns)
adapt_glm_rejs <- glm_ns$nrejs[seq_along(Q_LEVELS)]
```

### 8d. AdaPT-GMM (p-values)

```{r adapt-gmm-p}
gmm_ns_p         <- adapt_gmm(X, pvals = p, alphas = Q_LEVELS,
                               beta_formulas = formula_ns)
adapt_gmm_p_rejs <- gmm_ns_p$nrejs[seq_along(Q_LEVELS)]
```

### 8e. AdaPT-GMM (z-statistics)

```{r adapt-gmm-z}
gmm_ns_z         <- adapt_gmm(X, z = z, alphas = Q_LEVELS,
                               beta_formulas = formula_ns, testing = "two_sided")
adapt_gmm_z_rejs <- gmm_ns_z$nrejs[seq_along(Q_LEVELS)]
```

### 8f. OrderShapeEM

```{r ordershape}
os_x1 <- OrderShapeEM(pvals = p, order.var = X[,1],
                      OrderShapeEM.control(trace = FALSE))
os_x2 <- OrderShapeEM(pvals = p, order.var = X[,2],
                      OrderShapeEM.control(trace = FALSE))
os_x3 <- OrderShapeEM(pvals = p, order.var = X[,3],
                      OrderShapeEM.control(trace = FALSE))
os_x1_rejs <- sapply(Q_LEVELS, function(q) sum(os_x1$fdr <= q))
os_x2_rejs <- sapply(Q_LEVELS, function(q) sum(os_x2$fdr <= q))
os_x3_rejs <- sapply(Q_LEVELS, function(q) sum(os_x3$fdr <= q))
```

### 8g. FDRreg

```{r fdrreg}
n_unique_X  <- vapply(X, function(col) length(unique(col[is.finite(col)])), integer(1))
varying_X   <- names(n_unique_X)[n_unique_X > 1]

if (length(varying_X)) {
  set.seed(SEED)
  fdr_theo    <- FDRreg::FDRreg(z, as.matrix(X[, varying_X, drop = FALSE]),
                                nulltype = "theoretical")
  fdrreg_rejs <- sapply(Q_LEVELS, function(q) sum(fdr_theo$FDR <= q))
} else {
  fdr_theo    <- NULL
  fdrreg_rejs <- rep(NA_integer_, length(Q_LEVELS))
}
```

### 8h. IHW

```{r ihw, message=FALSE}
ihw_x1_rejs <- compute_ihw_rejs(p, X[,1], "xA", Q_LEVELS)
ihw_x2_rejs <- compute_ihw_rejs(p, X[,2], "xB", Q_LEVELS)
ihw_x3_rejs <- compute_ihw_rejs(p, X[,3], "xC", Q_LEVELS)
```


---

## 9. Main Figure - Number of Discoveries vs FDR Threshold

```{r main-figure, fig.width=10, fig.height=5}
LLM_fam <- data.frame(
  q               = Q_LEVELS,
  BH              = bh_rejs,
  chai            = chai_rejs,
  adapt_glm       = adapt_glm_rejs,
  adapt_gmm_p     = adapt_gmm_p_rejs,
  adapt_gmm_z     = adapt_gmm_z_rejs,
  OrderShapeEM_x1 = os_x1_rejs,
  # OrderShapeEM_x2 = os_x2_rejs,
  # OrderShapeEM_x3 = os_x3_rejs,
  FDRreg          = fdrreg_rejs,
  IHW_x1          = ihw_x1_rejs,
  IHW_x2          = ihw_x2_rejs,
  IHW_x3          = ihw_x3_rejs
)

LLM_fam_long <- LLM_fam %>%
  pivot_longer(
    cols = -q,
    names_to = "Method",
    values_to = "Rejections"
  ) %>%
  mutate(Method = as.character(Method))
```


```{r fig1}
source("consistant_color.R")
plot_rejections_vs_q(LLM_fam_long,
                     x_breaks = sort(unique(LLM_fam_long$q)))

```

---

## Session Info

```{r session-info}
sessionInfo()
```
