---
title: "Simulation_S2"
author: "Ziyi Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

## Overview

This document reproduces the Supplementary Simulation S2, which is based on the one-sided z-statistic setting 
with two-dimensional X as the side covariates (auxiliary information).

**Working directory:** Knit this file from (`chai_project/scripts/`).

---

## 1. Install Dependencies

Run once in a fresh R session before knitting:

```{r install, eval=FALSE}
# CRAN + Bioconductor + GitHub dependencies
source("install_packages.R")

# chai package (not on CRAN)
remotes::install_github("ziyiwang726/chai")
```

---

## 2. Load Packages

```{r packages, include=TRUE, warning=FALSE}
suppressPackageStartupMessages({
  library(chai)
  library(knitr)
  library(mvtnorm)
  library(mixtools)
  library(KScorrect)
  library(adaptMT)
  library(AdaPTGMM)
  library(splines2)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(coin)
  library(ggrepel)
  library(IHW)
  library(FDRreg)
  library(OrderShapeEM)
  library(patchwork)
  library(ggrepel)
  library(stringr)
})
```


---

## 3. Simulation generating and fit the models

**Note: **

This chunk may be very time-consuming. Specifically, for each seed (100 randomly generated seeds in total), one simulation dataset was generated and then analyzed using \code{chai} and each benchmark method. For every method, we recorded the number of selections, FDP, and power at all $q$-level from 0.01 to 0.10, together with the corresponding seed. 

By default, the code reads the previously generated RDS file to save time. If you want to re-generate the results from scratch, you can change the path of `cache_file` to let R cannot find the corresponding RDS file so that R starts by re-running the simulation.

```{r, warning=FALSE}
set.seed(1)
seeds <- c(sample(1:10000, 99), 123)   # Add seed 123 for checking purpose

n = 1000
n0 <- 950
n1 <- 25

# Ground truth for alternative
gt <- seq((n0+1), n)

q_grid <- seq(0.01, 0.1, by = 0.01)
alphas <- q_grid

results_list <- list()


cache_file <- "../data/chai_env/Simulation_S2_results.rds"

if (file.exists(cache_file)) {
  # if the file exists, load it
  results_s2 <- readRDS(cache_file)
} else {
  # if not, run the simulation
  for (s in seq_along(seeds)) {
    message("Running seed ", s, " / ", length(seeds), 
          "  (seed value = ", seeds[s], ")")

    set.seed(seeds[s])

    z0 <- rnorm(n0, mean = 0, sd = 1)
    x0 <- rnorm(n0, mean = 5, sd = 1)
    y0 <- rnorm(n0, mean = 1, sd = 1)
    
    z1 <- rnorm(n1, mean = 2, sd = 0.5)
    x1 <- rnorm(n1, mean = 4, sd = 1)
    y1 <- rnorm(n1, mean = 2, sd = 1)
    
    z2 <- rnorm(n1, mean = 3, sd = 0.5)
    x2 <- rnorm(n1, mean = 3, sd = 1)
    y2 <- rnorm(n1, mean = 3, sd = 1)
    
    z <- c(z0, z1, z2)
    p <- pnorm(z, lower.tail = FALSE) # One-sided

    x <- c(x0, x1, x2)
    y <- c(y0, y1, y2)
    
    X <- cbind(x, y)
  
    # Common values
    x_df <- data.frame(X)
    formula_ns <- list("~ns(x, df = 2) + ns(y, df = 2)", 
                       "~ns(x, df = 4) + ns(y, df = 4)",
                       "~ns(x, df = 6) + ns(y, df = 6)")

    # Apply models and collect performance
    perf_a <- data.frame(
      Seed    = integer(),
      q       = numeric(),
      Method  = character(),
      Selected  = numeric(),
      FDP       = numeric(),
      Power     = numeric(),
      stringsAsFactors = FALSE
    )
    
        # ------------------------------------------------------------------
    # 1. chai
    # ------------------------------------------------------------------
    set.seed(seeds[s])
    res_chai <- chai(z, X, B = 100)

    for (qq in seq_along(q_grid)) {
        select <- clfdrselect(res_chai, q = q_grid[qq])
        p1  <- performance(gt, select, z = res_chai$z,
                           bh_level = q_grid[qq], one.sided = TRUE)
        # chai
        perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "chai", p1[[4]], p1[[1]], p1[[2]])
        # BH
        perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "BH",   p1[[13]], p1[[8]], p1[[9]])
      }
    
    # ------------------------------------------------------------------
    # 2. adapt_gmm (z)
    # ------------------------------------------------------------------
    set.seed(seeds[s])
    res_gmm_z <- adapt_gmm(
      x_df, z = z, alphas = q_grid,
      beta_formulas = formula_ns, testing = "two_sided"
    )

    for (qq in seq_along(q_grid)) {
        p2  <- performance(gt, res_gmm_z$rejs[[qq]], z = z,
                           bh_level = q_grid[qq], one.sided = TRUE)
        perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "adapt_gmm_z", p2[[4]], p2[[1]], p2[[2]])
    }
    
    # ------------------------------------------------------------------
    # 3. adapt_gmm (p)
    # ------------------------------------------------------------------
    set.seed(seeds[s])
    res_gmm_p <- adapt_gmm(
      x_df, pvals = p, alphas = q_grid,
      beta_formulas = formula_ns, testing = "two_sided"
    )
 
    for (qq in seq_along(q_grid)) {
        # idx <- match(qq, alphas)
        p3  <- performance(gt, res_gmm_p$rejs[[qq]], z = z,
                           bh_level = q_grid[qq], one.sided = TRUE)
        perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "adapt_gmm_p", p3[[4]], p3[[1]], p3[[2]])
    }
    
    # ------------------------------------------------------------------
    # 4. adapt_glm
    # ------------------------------------------------------------------
    set.seed(seeds[s])
    res_glm <- adapt_glm(
      x_df, pvals = p, alphas = q_grid,
      pi_formulas = formula_ns, mu_formulas = formula_ns
    )

    empty <- is.null(res_glm) || (!is.list(res_glm$rejs))
    for (qq in seq_along(q_grid)) {
      if (!empty && lengths(res_glm$rejs)[qq] > 0L){
       p4 <- performance(gt, res_glm$rejs[[qq]], z = z,
                          bh_level = q_grid[qq], one.sided = TRUE)
       perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "adapt_glm", p4[[4]], p4[[1]], p4[[2]])
       
      } else {
        perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "adapt_glm", 0, 0, 0)
      }
    }
  
    # ------------------------------------------------------------------
    # 5. OrderShapeEM
    # ------------------------------------------------------------------
    # First x
    set.seed(seeds[s])
    res_osem_x1 <- NULL
    
    res_osem_x1 <- tryCatch(
      OrderShapeEM(pvals = p, order.var = X[, 1], OrderShapeEM.control(trace = TRUE)),
      error = function(e) {
        message(paste("OrderShapeEM_x1 failed at seed", seeds[s], ":", e$message))
        NULL
      }
    )

    
    if (!is.null(res_osem_x1)) {
      for (qq in seq_along(q_grid)) {
        p5_1 <- performance(gt, which(res_osem_x1$fdr <= q_grid[qq]), z = z, 
                            bh_level = q_grid[qq], one.sided = TRUE)
        
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x1",
                                       p5_1[[4]], p5_1[[1]], p5_1[[2]])
      }
    } else {
      for (qq in seq_along(q_grid)) {
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x1",
                                       NA, NA, NA)
      }
    }
  
    # Second x
    set.seed(seeds[s])
    res_osem_x2 <- NULL
    
    res_osem_x2 <- tryCatch(
      OrderShapeEM(pvals = p, order.var = X[, 2], OrderShapeEM.control(trace = TRUE)),
      error = function(e) {
        message(paste("OrderShapeEM_x2 failed at seed", seeds[s], ":", e$message))
        NULL
      }
    )
    
    if (!is.null(res_osem_x2)) {
      for (qq in seq_along(q_grid)) {
        p5_2 <- performance(gt, which(res_osem_x2$fdr <= q_grid[qq]), z = z, 
                            bh_level = q_grid[qq], one.sided = TRUE)
        
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x2",
                                       p5_2[[4]], p5_2[[1]], p5_2[[2]])
      }
    } else {
      for (qq in seq_along(q_grid)) {
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x2",
                                       NA, NA, NA)
      }
    }
  
    # ------------------------------------------------------------------
    # 6. FDRreg (theoretical null)
    # ------------------------------------------------------------------
    set.seed(seeds[s])
    res_fdrreg <- FDRreg(z, as.matrix(X), nulltype = "theoretical")

    for (qq in seq_along(q_grid)) {
      p6 <- performance(gt, which(res_fdrreg$FDR <= q_grid[qq]), z = z,
                        bh_level = q_grid[qq], one.sided = TRUE)
      perf_a[nrow(perf_a)+1,] <- list(seeds[s], q_grid[qq], "FDRreg", p6[[4]], p6[[1]], p6[[2]])
    }
    
    # ------------------------------------------------------------------
    # 7. IHW
    # ------------------------------------------------------------------
    # First x
    set.seed(seeds[s])
  # IHW: first x
    res_ihw_x1_list <- vector("list", length(q_grid))
    
    for (qq in seq_along(q_grid)) {
      res_ihw_x1_list[[qq]] <- tryCatch(
        ihw(pvalues    = p, covariates = X[, 1], alpha = q_grid[qq], nbins = 5),
        error = function(e) {
          message(paste("IHW_x1 failed at seed", seeds[s], "q =", q_grid[qq], ":", e$message))
          NULL
        }
      )
    }

    
    for (qq in seq_along(q_grid)) {
      if (!is.null(res_ihw_x1_list[[qq]])) {
        p7_1 <- performance(gt, which(adj_pvalues(res_ihw_x1_list[[qq]]) <= q_grid[qq]),
                            z = z, bh_level = q_grid[qq], one.sided = TRUE)
        
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x1",
                                       p7_1[[4]], p7_1[[1]], p7_1[[2]])
      } else {
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x1",
                                       NA, NA, NA)
      }
    }
  
    # Second x
    set.seed(seeds[s])
    res_ihw_x2_list <- vector("list", length(q_grid))
    
    for (qq in seq_along(q_grid)) {
      res_ihw_x2_list[[qq]] <- tryCatch(
        ihw(pvalues = p, covariates = X[, 2], alpha = q_grid[qq], nbins = 5),
        error = function(e) {
          message(paste("IHW_x2 failed at seed", seeds[s], "q =", q_grid[qq], ":", e$message))
          NULL
        }
      )
    }


    for (qq in seq_along(q_grid)) {
      if (!is.null(res_ihw_x2_list[[qq]])) {
        p7_2 <- performance(gt, which(adj_pvalues(res_ihw_x2_list[[qq]]) <= q_grid[qq]), 
                            z = z, bh_level = q_grid[qq], one.sided = TRUE)
        
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x2",
                                       p7_2[[4]], p7_2[[1]], p7_2[[2]])
      } else {
        perf_a[nrow(perf_a) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x2",
                                       NA, NA, NA)
      }
    }
    
    
    # Get the results
    results_list[[length(results_list) + 1]] <- perf_a

    # Clean up memory
    gc()
  } # end seeds loop

  # Combine and calculate average
  results_s2 <- dplyr::bind_rows(results_list) |>
    dplyr::mutate(
      Selected = as.numeric(Selected),
      FDP      = as.numeric(FDP),
      Power    = as.numeric(Power)
    )

  # Save to disk
  saveRDS(results_s2, cache_file)
}

```


---

## 4. Calculate the average across 100 repeats.

```{r}
summary_s2 <- results_s2 %>%
  group_by(Method, q) %>%
  summarise(
    mean_Selected = mean(Selected),
    mean_FDP      = mean(FDP),
    mean_Power    = mean(Power),
    .groups = "drop"
  )

summary_s2 %>% filter(q == 0.05)


sim_s2_long <- summary_s2 %>%
  mutate(
    mean_FDP   = ifelse(is.nan(mean_FDP), NA, mean_FDP),
    mean_Power = ifelse(is.nan(mean_Power), NA, mean_Power)
  ) %>%
  pivot_longer(cols = c(mean_FDP, mean_Power),
               names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = recode(Metric,
                         mean_FDP = "FDP",
                         mean_Power = "Power"))
```


---

## 5. Main Figure

### 5a. mean FDP across 100 repeats vs. FDR Threshold
### 5b. mean Power across 100 repeats vs. FDR Threshold


```{r}
source("consistant_color.R")

p_fdp   <- plot_metric(sim_s2_long, "FDP", "Mean FDP", "FDP vs q", fdp_y = 0.12)
p_power <- plot_metric(sim_s2_long, "Power", "Mean Power", "Power vs q")

combo <- p_fdp + p_power

```


```{r, fig.width=14, fig.height=5, out.width='100%', warning=FALSE}
combo

# ggsave(
#   filename = "../figures/Sim_S2.png",
#   plot = combo,
#   width = 14,
#   height = 5,
#   dpi = 300
# )
```

---

## Session Info

```{r session-info}
sessionInfo()
```

