---
title: "Simulation_2"
author: "Ziyi Wang"
date: "2026-03-23"
output: html_document
---
# Simulation 2 - Two dimensional x 
sim z

### Load Packages/Functions

```{r setup, include=TRUE, warning=FALSE}
suppressPackageStartupMessages({
  library(knitr)
  library(mvtnorm)
  library(mixtools)
  library(devtools)
  library(KScorrect)
  library(adaptMT)
  library(AdaPTGMM)
  library(splines2)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(coin)
  library(ggrepel)
  library(IHW)
  library(FDRreg)
  library(OrderShapeEM)
  library(chai)
})

```

### Simulation generating and fit the models

Note: this chunk is time-consuming. Specifically, for each seed (100 randomly generated seeds in total), one simulation dataset was generated and then analyzed using \code{chai} and each benchmark method. For every method, we recorded the number of selections, FDP, and power at all $q$-level from 0.01 to 0.10, together with the corresponding seed. 

You can directly run next chuck to load all the results.

```{r, eval=FALSE}
set.seed(1)
# seeds <- sample(1:10000, 99)
seeds <- sample(1:10000, 9)

seeds <- c(seeds, 123)    # Add seed 123 for checking purpose

q_grid <- seq(0.01, 0.1, by = 0.01)

results_list <- list()   # saving results
runtime_list <- list()   # saving computational time

add_runtime <- function(runtime_df, seed, method, fit_time, success) {
  runtime_df[nrow(runtime_df) + 1, ] <- list(seed, method, fit_time, success)
  runtime_df
}

for (s in seq_along(seeds)) {
  set.seed(seeds[s])
  
  n = 1000
  n0 <- 950
  n1 <- 25
  
  z0 <- rnorm(n0, mean = 0, sd = 1)
  x0 <- rnorm(n0, mean = 0, sd = 1)
  y0 <- rnorm(n0, mean = 0, sd = 1)
  
  z1 <- rnorm(n1, mean = 2, sd = 0.5)
  x1 <- rnorm(n1, mean = 2, sd = 1)
  y1 <- rnorm(n1, mean = 3, sd = 1)
  
  z2 <- rnorm(n1, mean = -2, sd = 0.5)
  x2 <- rnorm(n1, mean = -2, sd = 1)
  y2 <- rnorm(n1, mean = -3, sd = 1)
  
  z <- c(z0, z1, z2)
  x <- c(x0, x1, x2)
  y <- c(y0, y1, y2)
  
  X <- cbind(x, y)

  gt <- seq((n0+1), n) # Ground truth for alternative
  
  x_df <- data.frame(X)
  formula_ns <- list("~ns(x, df = 2) + ns(y, df = 2)", 
                     "~ns(x, df = 4) + ns(y, df = 4)", 
                     "~ns(x, df = 6) + ns(y, df = 6)")
  # p <- pnorm(z, lower.tail = FALSE)
  p <- 2 * (1 - pnorm(abs(z)))

  # Apply models and collect performance
  perf <- data.frame(
    Seed = integer(),
    q    = numeric(),
    Method   = character(),
    Selected = numeric(),
    FDP      = numeric(),
    Power    = numeric(),
    stringsAsFactors = FALSE
  )
  
  runtime <- data.frame(
    Seed = integer(),
    Method = character(),
    FitTime_sec = numeric(),
    Success = logical(),
    stringsAsFactors = FALSE
  )

  # 1. CHAI
  fit_time <- system.time({
    res_chai <- chai(z, X, B = 100)
  })["elapsed"]
  
  runtime <- add_runtime(runtime, seeds[s], "chai", as.numeric(fit_time), !is.null(res_chai))
  
  for (qq in seq_along(q_grid)) {
      select <- clfdrselect(res_chai, q = q_grid[qq])
      p1  <- performance(gt, select, z = res_chai$z,
                         bh_level = q_grid[qq], one.sided = FALSE)
      # chai
      perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "chai", p1[[4]], p1[[1]], p1[[2]])
      # BH
      perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "BH",   p1[[13]], p1[[8]], p1[[9]])
    }
  
  # 2. adapt_glm
  fit_time <- system.time({
    res_glm <- adapt_glm(
      x_df, pvals = p, alphas = q_grid,
      pi_formulas = formula_ns, mu_formulas = formula_ns
    )
  })["elapsed"]
  
  runtime <- add_runtime(runtime, seeds[s], "adapt_glm", as.numeric(fit_time), !is.null(res_glm))
  
  empty <- is.null(res_glm) || (!is.list(res_glm$rejs))
  for (qq in seq_along(q_grid)) {
    if (!empty && lengths(res_glm$rejs)[qq] > 0L){
     p2 <- performance(gt, res_glm$rejs[[qq]], z = z,
                        bh_level = q_grid[qq], one.sided = FALSE)
     perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "adapt_glm", p2[[4]], p2[[1]], p2[[2]])
     
    } else {
      perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "adapt_glm", 0, NA, NA)
    }
  }
  
  # 3. adapt_gmm_z
  fit_time <- system.time({
    res_gmm_z <- adapt_gmm(
      x_df, z = z, alphas = q_grid,
      beta_formulas = formula_ns, testing = "two_sided"
    )
  })["elapsed"]
  
  runtime <- add_runtime(runtime, seeds[s], "adapt_gmm_z", as.numeric(fit_time), !is.null(res_gmm_z))
  
  for (qq in seq_along(q_grid)) {
      # idx <- match(qq, alphas)  # 0.05 -> 5, 0.06 -> 6, ..., 0.10 -> 10
      p3  <- performance(gt, res_gmm_z$rejs[[qq]], z = z,
                         bh_level = q_grid[qq], one.sided = FALSE)
      perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "adapt_gmm_z", p3[[4]], p3[[1]], p3[[2]])
    }
  
  # 4. adapt_gmm_p
  fit_time <- system.time({
    res_gmm_p <- adapt_gmm(
      x_df, pvals = p, alphas = q_grid,
      beta_formulas = formula_ns, testing = "two_sided"
    )
  })["elapsed"]
  
  runtime <- add_runtime(runtime, seeds[s], "adapt_gmm_p", as.numeric(fit_time), !is.null(res_gmm_p))

  for (qq in seq_along(q_grid)) {
      # idx <- match(qq, alphas)
      p4  <- performance(gt, res_gmm_p$rejs[[qq]], z = z,
                         bh_level = q_grid[qq], one.sided = FALSE)
      perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "adapt_gmm_p", p4[[4]], p4[[1]], p4[[2]])
    }


  # 5. OrderShapeEM
  # First x
  res_osem_x1 <- NULL
  
  fit_time_x1 <- system.time({
    res_osem_x1 <- tryCatch(
      OrderShapeEM(pvals = p, order.var = X[, 1], OrderShapeEM.control(trace = TRUE)),
      error = function(e) {
        message(paste("OrderShapeEM_x1 failed at seed", seeds[s], ":", e$message))
        NULL
      }
    )
  })["elapsed"]
  
  runtime <- add_runtime(
    runtime_df = runtime,
    seed       = seeds[s],
    method     = "OrderShapeEM_x1",
    fit_time   = as.numeric(fit_time_x1),
    success    = !is.null(res_osem_x1)
  )
  
  if (!is.null(res_osem_x1)) {
    for (qq in seq_along(q_grid)) {
      p5_1 <- performance(gt, which(res_osem_x1$fdr <= q_grid[qq]), z = z, 
                          bh_level = q_grid[qq], one.sided = FALSE)
      
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x1",
                                     p5_1[[4]], p5_1[[1]], p5_1[[2]])
    }
  } else {
    for (qq in seq_along(q_grid)) {
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x1",
                                     NA, NA, NA)
    }
  }

  # Second x
  res_osem_x2 <- NULL
  
  fit_time_x2 <- system.time({
    res_osem_x2 <- tryCatch(
      OrderShapeEM(pvals = p, order.var = X[, 2], OrderShapeEM.control(trace = TRUE)),
      error = function(e) {
        message(paste("OrderShapeEM_x2 failed at seed", seeds[s], ":", e$message))
        NULL
      }
    )
  })["elapsed"]
  
  runtime <- add_runtime(
    runtime_df = runtime,
    seed       = seeds[s],
    method     = "OrderShapeEM_x2",
    fit_time   = as.numeric(fit_time_x2),
    success    = !is.null(res_osem_x2)
  )
  
  if (!is.null(res_osem_x2)) {
    for (qq in seq_along(q_grid)) {
      p5_2 <- performance(gt, which(res_osem_x2$fdr <= q_grid[qq]), z = z, 
                          bh_level = q_grid[qq], one.sided = FALSE)
      
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x2",
                                     p5_2[[4]], p5_2[[1]], p5_2[[2]])
    }
  } else {
    for (qq in seq_along(q_grid)) {
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "OrderShapeEM_x2",
                                     NA, NA, NA)
    }
  }

  # 6. FDRreg
  fit_time <- system.time({
    res_fdrreg <- FDRreg(z, as.matrix(X), nulltype = "theoretical")
  })["elapsed"]
  
  runtime <- add_runtime(runtime, seeds[s], "FDRreg", as.numeric(fit_time), !is.null(res_fdrreg))
  
  for (qq in seq_along(q_grid)) {
    p6 <- performance(gt, which(res_fdrreg$FDR <= q_grid[qq]), z = z,
                      bh_level = q_grid[qq], one.sided = FALSE)
    perf[nrow(perf)+1,] <- list(seeds[s], q_grid[qq], "FDRreg", p6[[4]], p6[[1]], p6[[2]])
  }
  
  # 7. IHW
  # First x
# IHW: first x
  res_ihw_x1_list <- vector("list", length(q_grid))
  
  fit_time_ihw_x1 <- system.time({
    for (qq in seq_along(q_grid)) {
      res_ihw_x1_list[[qq]] <- tryCatch(
        ihw(pvalues    = p, covariates = X[, 1], alpha = q_grid[qq], nbins = 5),
        error = function(e) {
          message(paste("IHW_x1 failed at seed", seeds[s], "q =", q_grid[qq], ":", e$message))
          NULL
        }
      )
    }
  })["elapsed"]
  
  runtime <- add_runtime(
    runtime_df = runtime,
    seed       = seeds[s],
    method     = "IHW_x1",
    fit_time   = as.numeric(fit_time_ihw_x1),
    success    = all(!sapply(res_ihw_x1_list, is.null))
  )
  
  for (qq in seq_along(q_grid)) {
    if (!is.null(res_ihw_x1_list[[qq]])) {
      p7_1 <- performance(gt, which(adj_pvalues(res_ihw_x1_list[[qq]]) <= q_grid[qq]),
                          z = z, bh_level = q_grid[qq], one.sided = FALSE)
      
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x1",
                                     p7_1[[4]], p7_1[[1]], p7_1[[2]])
    } else {
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x1",
                                     NA, NA, NA)
    }
  }

  # Second x
  res_ihw_x2_list <- vector("list", length(q_grid))
  
  fit_time_ihw_x2 <- system.time({
    for (qq in seq_along(q_grid)) {
      res_ihw_x2_list[[qq]] <- tryCatch(
        ihw(pvalues = p, covariates = X[, 2], alpha = q_grid[qq], nbins = 5),
        error = function(e) {
          message(paste("IHW_x2 failed at seed", seeds[s], "q =", q_grid[qq], ":", e$message))
          NULL
        }
      )
    }
  })["elapsed"]
  
  runtime <- add_runtime(
    runtime_df = runtime,
    seed       = seeds[s],
    method     = "IHW_x2",
    fit_time   = as.numeric(fit_time_ihw_x2),
    success    = all(!sapply(res_ihw_x2_list, is.null))
  )
  
  for (qq in seq_along(q_grid)) {
    if (!is.null(res_ihw_x2_list[[qq]])) {
      p7_2 <- performance(gt, which(adj_pvalues(res_ihw_x2_list[[qq]]) <= q_grid[qq]), 
                          z = z, bh_level = q_grid[qq], one.sided = FALSE)
      
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x2",
                                     p7_2[[4]], p7_2[[1]], p7_2[[2]])
    } else {
      perf[nrow(perf) + 1, ] <- list(seeds[s], q_grid[qq], "IHW_x2",
                                     NA, NA, NA)
    }
  }
  
  
  # Get the results
  results_list[[length(results_list) + 1]] <- perf
  runtime_list[[length(runtime_list) + 1]] <- runtime
}

# Combine and calculate average
results_2 <- dplyr::bind_rows(results_list) |>
  mutate(
    Selected = as.numeric(Selected),
    FDP      = as.numeric(FDP),
    Power    = as.numeric(Power)
  )

runtime_2 <- dplyr::bind_rows(runtime_list) |>
  dplyr::mutate(FitTime_sec = as.numeric(FitTime_sec))

runtime_summary <- runtime_2 |>
  dplyr::group_by(Method) |>
  dplyr::summarise(
    mean_fit_time   = mean(FitTime_sec, na.rm = TRUE),
    sd_fit_time     = sd(FitTime_sec, na.rm = TRUE),
    median_fit_time = median(FitTime_sec, na.rm = TRUE),
    .groups = "drop"
  )

```

#### Load simulation 2's results 
Load simulation 2 results and calculate the average across 100 repeats.

```{r}
results_2 <- readRDS("./data/chai_env/simZ_all_results_with_zero.rds")

summary_2 <- results_2 %>%
  group_by(Method, q) %>%
  summarise(
    mean_Selected = mean(Selected),
    mean_FDP      = mean(FDP),
    mean_Power    = mean(Power),
    .groups = "drop"
  )

summary_2 %>% filter(q == 0.05)


simZ_long <- summary_2 %>%
  mutate(
    mean_FDP   = ifelse(is.nan(mean_FDP), NA, mean_FDP),
    mean_Power = ifelse(is.nan(mean_Power), NA, mean_Power)
  ) %>%
  pivot_longer(cols = c(mean_FDP, mean_Power),
               names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = recode(Metric,
                         mean_FDP = "FDP",
                         mean_Power = "Power"))
```

### Define the plot functions to ensure consistent color assignments across different methods.
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# ---------- color helpers ----------
blend_color <- function(col, target = "white", amount = 0.4) {
  a <- grDevices::col2rgb(col) / 255
  b <- grDevices::col2rgb(target) / 255
  out <- (1 - amount) * a + amount * b
  grDevices::rgb(out[1,], out[2,], out[3,])
}

variant_shades <- function(base_col, K = 12) {
  grDevices::colorRampPalette(c(
    blend_color(base_col, "white", 0.55),
    base_col,
    blend_color(base_col, "black", 0.25)
  ))(K)
}

make_method_palette <- function(methods, K = 12) {
  fixed <- c(
    chai        = "red",
    adapt_glm   = "#FF7F0E",
    adapt_gmm_z = "#2CA02C",
    adapt_gmm_p = "#9467BD",
    FDRreg      = "#8C564B",
    BH          = "black"
  )

  family_base <- c(
    OrderShapeEM = "#1F77B4",  # blue family
    IHW          = "#17BECF"   # teal family
  )

  ord_shades <- variant_shades(family_base["OrderShapeEM"], K = K)
  ihw_shades <- variant_shades(family_base["IHW"],          K = K)

  methods <- unique(as.character(methods))
  pal <- setNames(rep("#7F7F7F", length(methods)), methods)  # fallback grey

  for (m in methods) {
    if (m %in% names(fixed)) {
      pal[m] <- fixed[m]
      next
    }

    base <- sub("_x\\d+$", "", m)
    idx  <- suppressWarnings(as.integer(sub(".*_x", "", m)))

    if (base %in% names(family_base)) {
      if (is.na(idx)) {
        pal[m] <- family_base[base]  # e.g. "OrderShapeEM" with no suffix
      } else {
        idx <- max(1L, min(idx, K))
        pal[m] <- if (base == "OrderShapeEM") ord_shades[idx] else ihw_shades[idx]
      }
    }
  }

  pal
}

method_order <- function(methods) {
  methods <- unique(as.character(methods))
  core <- c("chai","adapt_glm","adapt_gmm_z","adapt_gmm_p","FDRreg")

  sort_family <- function(base) {
    fam <- methods[grepl(paste0("^", base, "(_x\\d+)?$"), methods)]
    base_present <- base %in% fam
    vars <- fam[grepl("_x\\d+$", fam)]
    if (length(vars)) {
      idx <- suppressWarnings(as.integer(sub(".*_x", "", vars)))
      vars <- vars[order(idx)]
    }
    c(if (base_present) base else character(0), vars)
  }

  out <- c(core, sort_family("OrderShapeEM"), sort_family("IHW"), "BH")
  out <- unique(out)
  out[out %in% methods]
}

# ---------- plotting function (labels, no legend) ----------
library(ggrepel)
library(stringr)

plot_metric <- function(long_table, metric_name, 
                        ylab, title, K_variants = 12, 
                        fdp_y = 0.2, method_font = 5.5){
  
  dat <- dplyr::filter(long_table, Metric == metric_name) %>%
    mutate(Method = as.character(Method))

  ord <- method_order(dat$Method)
  dat <- dat %>% mutate(Method = factor(Method, levels = ord))
  pal <- make_method_palette(levels(dat$Method), K = K_variants)

  # endpoints for labels
  endpoints <- dat %>%
    group_by(Method) %>%
    filter(q == max(q, na.rm = TRUE)) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    mutate(
      # option A: wrap long names
      label = as.character(Method)
      # label = str_wrap(label, width = 12)
    )

  x_rng <- range(dat$q, na.rm = TRUE)
  x_nudge <- diff(x_rng) * 0.03  # more room to the right

  p <- ggplot(dat, aes(x = q, y = Value, group = Method, color = Method)) +
    geom_line(linewidth = 0.9, na.rm = TRUE) +
    geom_point(size = 1.6, na.rm = TRUE) +

    # repel labels to avoid overlap
    ggrepel::geom_text_repel(
      data = endpoints,
      aes(label = label),
      size = method_font,
      nudge_x = x_nudge,
      hjust = 0,
      direction = "y",
      box.padding = 0.25,
      point.padding = 0.15,
      segment.size = 0.25,
      min.segment.length = 0,
      max.overlaps = Inf,
      seed = 1,
      show.legend = FALSE
    ) +

    scale_color_manual(values = pal, breaks = ord) +
    scale_x_continuous(
      breaks = seq(0.01, 0.10, 0.01),
      expand = expansion(mult = c(0.02, 0.18))  # add right padding for labels
    ) +
    coord_cartesian(
      ylim = c(0, 1),
      clip = "off"   # <-- critical: do not clip labels
    ) +
    labs(x = "q level", y = ylab, title = title) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.margin = margin(5.5, 60, 5.5, 5.5),  # <-- more right margin
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14),
      axis.text.x  = element_text(size = 12),
      axis.text.y  = element_text(size = 12)
    )

  if (metric_name == "FDP") {
    p <- p +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 1.2) +
      coord_cartesian(ylim = c(0, fdp_y))
  }

  p
}
```

### Visualization

```{r}
p_fdp   <- plot_metric(simZ_long, "FDP", "Mean FDP", "FDP vs q", fdp_y = 0.2)
p_power <- plot_metric(simZ_long, "Power", "Mean Power", "Power vs q")

library(patchwork)
combo <- p_fdp + p_power # + plot_annotation(title = "Method Performance vs q")

```

```{r, fig.width=14, fig.height=5, out.width='100%', warning=FALSE}
combo
```
